---
title: "PLIER Utils proof-of-concept"
output:
  html_notebook: default
  pdf_document: default
---
**J. Taroni 2018** 

In this notebook, I aim to demonstrate the validity of implementation for a 
subset of PLIER custom functions. Specifically, checking the operations for:

* Training a PLIER model on a new dataset `PLIERNewData()`
* Applying a previously computed PLIER to a new dataset to get the LV x sample
  matrix (B) `GetNewDataB()`
* Reconstruction of input gene expression data with a PLIER model 
  `GetReconstructedExprs()` and the evaluation function 
  `GetReconstructionCorrelation()`

Here, I use the **NARES dataset** for convenience due to its relatively small 
size (n = 77). 

#### Load libraries and custom functions
```{r}
library(AnnotationDbi)
library(PLIER)

source(file.path("util", "plier_util.R"))
```

```{r}
# plots directory specifically for this notebook
dir.create(file.path("plots", "01"), recursive = TRUE,
           showWarnings = FALSE)
```


#### NARES expression data

```{r}
# Read in PCL
nares.data <- readr::read_tsv(file.path("data", "expression_data", 
                                         "NARES_SCANfast_ComBat.pcl"),
                              progress = FALSE)
```
Building a PLIER model requires HGNC symbol annotation, as this is what is
included in the prior information (e.g., pathways, genesets) that is included
in the package.

```{r}
symbol.obj <- org.Hs.eg.db::org.Hs.egSYMBOL
mapped.genes <- AnnotationDbi::mappedkeys(symbol.obj)
symbol.list <- as.list(symbol.obj[mapped.genes])
symbol.df <- as.data.frame(cbind(names(symbol.list), unlist(symbol.list)))
colnames(symbol.df) <- c("EntrezID", "GeneSymbol")

# get gene column name to match to facilitate use with dplyr
colnames(nares.data)[1] <- "EntrezID"

# matching types
symbol.df$EntrezID <- as.integer(as.character(symbol.df$EntrezID))

# inner join
annot.nares.data <- dplyr::inner_join(symbol.df, nares.data, by = "EntrezID")

# only leave the matrix with gene symbol as rownames
exprs.mat <- dplyr::select(annot.nares.data, -EntrezID)
rownames(exprs.mat) <- exprs.mat$GeneSymbol
exprs.mat <- as.matrix(dplyr::select(exprs.mat, -GeneSymbol))
```

#### Train PLIER model

`PLIERNewData` is a wrapper function for applying PLIER to a new dataset. 
Expression data is row-normalized for use with PLIER.
We use the following genesets that come with PLIER: `bloodCellMarkersIRISDMAP`, 
`svmMarkers`, and `canonicalPathways`.
See also the [PLIER vignette](https://github.com/wgmao/PLIER/blob/a2d4a2aa343f9ed4b9b945c04326bebd31533d4d/vignettes/vignette.pdf).

```{r}
nares.plier <- PLIERNewData(exprs.mat)
```

#### Apply PLIER model to "new" expression dataset

The `GetNewDataB` function will first row-normalize and reorder the "new" input 
gene expression data (`exprs.mat`), and then using a previously computed PLIER 
model (`plier.model`, specifically the gene loadings and the L2 constant), get
the new data into the PLIER model LV space. Here, we supply the NARES data as
the expression data and the PLIER model that has already been trained on the
same gene expression matrix.

```{r}
new.b.mat <- GetNewDataB(exprs.mat = exprs.mat,
                         plier.model = nares.plier)
# NARES B matrix from PLIERNewData
nares.b.mat <- nares.plier$B
```
We want to ensure that the two B matrices are the same.

```{r}
all.equal(nares.b.mat, new.b.mat)
```

#### Reconstruction of gene expression data

We reconstruct gene expression data from the gene loadings and LVs.

```{r}
nares.recon <- GetReconstructedExprs(z.matrix = as.matrix(nares.plier$Z),
                                     b.matrix = as.matrix(nares.plier$B))
```

Let's evaluate the reconstruction. 
We'll need the row-normalized NARES expression data (input) for comparison.

```{r}
nares.rownorm <- PLIER::rowNorm(exprs.mat)
nares.rownorm <- nares.rownorm[rownames(nares.recon), ]
```

Spearman correlation between input, row-normalized expression data and the
reconstructed data.
If correlation between the input and the reconstructed data is high, that
suggests that reconstruction is "successful." 
Given the different constraints in PLIER, we would not expect to perfectly 
(`rho = 1`) reconstruct the input data.
This particular evaluation will be _most useful_ when we look at applying a
trained PLIER model to a test dataset.
```{r}
recon.correlation <- GetReconstructionCorrelation(true.mat = nares.rownorm,
                                                  recon.mat = nares.recon)
```

Plot density
```{r}
# density plot
p <- ggplot2::ggplot(as.data.frame(recon.correlation), 
                ggplot2::aes(x = recon.correlation)) + 
      ggplot2::geom_density() +
      ggplot2::theme_bw() +
      ggplot2::labs(x = "Spearman Correlation",
                    title = "Input vs. PLIER reconstructed NARES data") +
      ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, 
                                                        face = "bold"))
p
```

```{r}
png.file <- file.path("plots", "01",
                      "NARES_reconstructed_data_correlation_density.png")
ggplot2::ggsave(filename = png.file, plot = p, width = 7, height = 5,
                units = "in")
```
