---
title: "Isolated immune cell populations: microarray data"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

[E-MTAB-2452](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-2452/): 
Sorted peripheral blood cells (CD4+ T cells, CD14+ monocytes, CD16+ neutrophils)
profiled on microarray from several autoimmune diseases.
This dataset is a good control for further investigating recount2 PLIER model 
(cross-platform) transfer learning.

## Functions and set up

```{r}
`%>%` <- dplyr::`%>%`
library(AnnotationDbi)
# custom functions
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "03")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "03")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

### `recount2` PLIER model

```{r}
plier.results <- readRDS(file.path("data", "recount2_PLIER_data", 
                                   "recount_PLIER_model.RDS"))
```

### Isolated cell subsets microarray data
```{r}
# read in E-MTAB-2452 expression data that has been processed with SCANfast
exprs.df <- readr::read_tsv(file.path("data", "expression_data",
                                      "E-MTAB-2452_hugene11st_SCANfast.pcl"))
colnames(exprs.df)[1] <- "EntrezID"
```

#### Conversion to gene symbol
```{r}
# remove trailing "_at" (result of using BrainArray) to get EntrezIDs
exprs.df$EntrezID <- gsub("_at", "", exprs.df$EntrezID)

# conversion to gene symbol
symbol.obj <- org.Hs.eg.db::org.Hs.egSYMBOL
mapped.genes <- AnnotationDbi::mappedkeys(symbol.obj)
symbol.list <- as.list(symbol.obj[mapped.genes])
symbol.df <- as.data.frame(cbind(names(symbol.list), unlist(symbol.list)))
colnames(symbol.df) <- c("EntrezID", "GeneSymbol")

# inner join
annot.data <- dplyr::inner_join(symbol.df, exprs.df, by = "EntrezID")

# remove objects only necessary for annotation
rm(mapped.genes, symbol.list, symbol.obj, symbol.df)

# write to file
gs.file <- file.path("data", "expression_data", 
                     "E-MTAB-2452_hugene11st_SCANfast_with_GeneSymbol.pcl")
readr::write_tsv(annot.data, path = gs.file)

# get as a matrix 
exprs.mat <- as.matrix(annot.data[, 3:ncol(annot.data)])
rownames(exprs.mat) <- annot.data$GeneSymbol
```

## Apply recount2 model

### B matrix
```{r}
iso.b.matrix <- GetNewDataB(exprs.mat = exprs.mat,
                            plier.model = plier.results)
head(iso.b.matrix[, 1:10])
```
```{r}
# write to file
iso.b.file <- file.path(results.dir, "E-MTAB-2452_B_matrix_recount2_model.txt")
write.table(iso.b.matrix, file = iso.b.file, sep = "\t", quote = FALSE)
```

### Heatmap

```{r}
# differences in cell type LVs?
indx.relevant.lv <- c(grep("CD4", rownames(iso.b.matrix)),
                      grep("Monocyte", rownames(iso.b.matrix)),
                      grep("Neutrophil", rownames(iso.b.matrix)))

# get row side color vector matching relevant cell type LVs
row.colors.vec <- c(rep("#1E90EF", sum(grepl("CD4", rownames(iso.b.matrix)))),
                    rep("#000000", 
                        sum(grepl("Monocyte", rownames(iso.b.matrix)))), 
                    rep("#32CD32",
                        sum(grepl("Neutrophil", rownames(iso.b.matrix)))))

# get col side color vector based on what cell type each sample is
col.colors.vec <- rep("white", ncol(iso.b.matrix))
col.colors.vec[grep("CD4", colnames(iso.b.matrix))] <- "#1E90EF"
col.colors.vec[grep("CD14", colnames(iso.b.matrix))] <- "#000000"
col.colors.vec[grep("CD16", colnames(iso.b.matrix))] <- "#32CD32"

# heatmap
gplots::heatmap.2(iso.b.matrix[indx.relevant.lv, ], 
                        trace = "none",
                        col = colorRampPalette(c("blue", "white", "red")),
                        cexCol = 0.5, cexRow = 0.6,
                        RowSideColors = row.colors.vec,
                        ColSideColors = col.colors.vec,
                        margins = c(6, 10))
legend("topright", 
       legend = c("CD4 T cell", "CD14, monocyte", "CD16, neutrophil"),
       col = c("#1E90EF", "#000000", "#32CD32"), cex = 0.5,
       lty = 1, lwd = 6)
```
```{r}
# save heatmap as pdf -- can not figure out another way around this, hm just
# returns a list
pdf(file.path(plot.dir, "E-MTAB-2452_recount_PLIER_cell_type_LVs_B.pdf"))
gplots::heatmap.2(iso.b.matrix[indx.relevant.lv, ], 
                        trace = "none",
                        col = colorRampPalette(c("blue", "white", "red")),
                        cexCol = 0.5, cexRow = 0.6,
                        RowSideColors = row.colors.vec,
                        ColSideColors = col.colors.vec,
                        margins = c(6, 10))
legend("topright", 
       legend = c("CD4 T cell", "CD14, monocyte", "CD16, neutrophil"),
       col = c("#1E90EF", "#000000", "#32CD32"), cex = 0.5,
       lty = 1, lwd = 6)
dev.off()
```

