---
title: "Isolated immune cell reconstruction"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---
**J. Taroni 2018**

In this notebook, we'll be evaluating reconstruction of [`E-MTAB-2452`](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-2452/) 
sorted peripheral blood cells (CD4+ T cells, CD14+ monocytes, CD16+ neutrophils)
profiled on microarray from several autoimmune diseases.
as compared to isolated immune cell data included in the recount2.
[`SRP045500`](https://trace.ncbi.nlm.nih.gov/Traces/sra/?study=SRP045500) is a 
comparison of 6 cell subsets in "immune-associated diseases."

## Functions and directory set up

```{r}
`%>%` <- dplyr::`%>%`
# custom functions
source(file.path("util", "plier_util.R"))
```
```{r}
# Noticed I needed this functionality repeatedly
ReconstructionEvalWrapper <- function(z.mat, b.mat, input.exprs,
                                      dataset.name, lv.type = "All") {
  # This function is a wrapper for reconstructing gene expression data given
  # gene loadings (Z) and LVs (B) from a PLIER model. It also takes the
  # input expression data so evaluations (Spearman correlation & MASE) can be
  # performed. Finally, it returns a tidy data.frame of these measures.
  # See also: the compare reconstructed to input section of util/plier_util.R
  #
  # Args:
  #   z.mat: a matrix of gene loadings from PLIER
  #   b.mat: a matrix of latent variables from PLIER
  #   input.exprs: gene expression matrix before reconstruction (input into 
  #               PLIER), used as the `true.mat` argument for the reconstruction
  #               evaluations (should be row-normalized, can use
  #               GetOrderedRowNorm to obtain this matrix)
  #   dataset.name: a string that indicates the dataset being reconstructed,
  #                 this will end up being the value in the `Dataset` column of
  #                 data.frame returned by this function
  #   lv.type: a string that indicates what LVs were use for reconstruction,
  #            this will end up being the value in the 
  #            `LVs used in reconstruction` column of the data.frame returned
  #
  # Returns:
  #   recon.eval.df: A data.frame with the following columns - Sample, MASE, 
  #                  Spearman correlation, LVs used in reconstruction, Dataset
  #
  
  ## Reconstruction
  recon.exprs <- GetReconstructedExprs(z.matrix = z.mat, 
                                       b.matrix = b.mat)
  
  ## Evaluations
  recon.error <- GetReconstructionMASE(true.mat = input.exprs,
                                       recon.mat = recon.exprs)
  recon.cor <- GetReconstructionCorrelation(true.mat = input.exprs,
                                            recon.mat = recon.exprs)
  
  ## get data.frame
  num.samples <- ncol(input.exprs)
  recon.eval.df <- as.data.frame(cbind(colnames(input.exprs), 
                                       recon.error, recon.cor, 
                                       rep(lv.type, num.samples), 
                                       rep(dataset.name, num.samples)))
  colnames(recon.eval.df) <- c("Sample", "MASE", "Spearman correlation",
                               "LVs used in reconstruction", "Dataset")
  recon.eval.df <- recon.eval.df %>%
                    dplyr::mutate(MASE = as.numeric(as.character(MASE)),
                    `Spearman correlation` = 
                      as.numeric(as.character(`Spearman correlation`)))
  
  return(recon.eval.df)
}
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "04")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "04")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data

### E-MTAB-2452

#### Expression data

```{r}
gs.file <- file.path("data", "expression_data", 
                     "E-MTAB-2452_hugene11st_SCANfast_with_GeneSymbol.pcl")
exprs.df <-readr::read_tsv(gs.file)
# drop gene identifiers
exprs.mat <- as.matrix(exprs.df[, 3:ncol(exprs.df)])
rownames(exprs.mat) <- exprs.df$GeneSymbol
rm(exprs.df)
```

#### B matrix

From recount2 model

```{r}
iso.b.file <- file.path("results", "03", 
                        "E-MTAB-2452_B_matrix_recount2_model.txt")
iso.b.matrix <- read.delim(iso.b.file)
```

### recount2

#### PLIER model

```{r}
plier.results <- readRDS(file.path("data", "recount2_PLIER_data", 
                                   "recount_PLIER_model.RDS"))
```

#### Input expression data

```{r}
# data that was prepped for use with PLIER
recount.list <- readRDS(file.path("data", "recount2_PLIER_data", 
                                  "recount_data_prep_PLIER.RDS"))
# input expression data from intermediate file
recount.input.exprs <- recount.list$rpkm.cm
rm(recount.list)
```

#### Evaluation of reconstruction

Spearman correlation, MASE

```{r}
eval.file <- file.path("results", "02", 
                       "recount2_recount2_model_recon_eval_df.tsv")
recount.eval.df <- readr::read_tsv(eval.file)
recount.eval.df$Dataset <- rep("recount2", nrow(recount.eval.df))
```

Identify the samples that belong to `SRP045500`, we'll duplicate these values
in `recount.eval.df` so we can plot them separately.

```{r}
srp.df <- recount.eval.df %>%
            dplyr::filter(grepl("SRP045500", recount.eval.df$Sample)) %>%
            dplyr::mutate(Dataset = "SRP045500")
recount.eval.df <- dplyr::bind_rows(recount.eval.df, srp.df)
```

## Reconstruction of E-MTAB-2452 with all LVs

```{r}
# recount2 model Z matrix
z.matrix <- plier.results$Z
```

### Reconstruction and evaluation

We need to obtain a (row-normalized) gene expression matrix as the "true"
expression matrix. 
We use this for evaluation with Spearman and MASE as in previous notebooks.
We'll use this for the other evaluations (e.g., pathway-associated LVs only) as 
well.

```{r}
iso.ord.rownorm <- GetOrderedRowNorm(exprs.mat = exprs.mat,
                                     plier.model = plier.results)  #recount2
```

```{r}
iso.recon.all.df <- ReconstructionEvalWrapper(z.mat = as.matrix(z.matrix),
                                              b.mat = as.matrix(iso.b.matrix),
                                              input.exprs = iso.ord.rownorm,
                                              lv.type = paste("All, n =", 
                                                              ncol(z.matrix)),
                                              dataset.name = "E-MTAB-2452")
```

```{r}
head(iso.recon.all.df)
```
### Plotting 

Add the isolated cell type reconstruction evaluation measures to the recount
data.frame

```{r}
eval.df <- dplyr::bind_rows(recount.eval.df, iso.recon.all.df)
rm(iso.recon.all.df)
```
#### Correlation
```{r}
dplyr::filter(eval.df, `LVs used in reconstruction` == 
                paste("All, n =", ncol(z.matrix))) %>%
  ggplot2::ggplot(ggplot2::aes(x = `Spearman correlation`,
                               group = Dataset, fill = Dataset)) +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::theme_bw() +
  ggplot2::scale_fill_manual(values = c("white", "gray50", "black")) +
  ggplot2::ggtitle(paste("All, n =", ncol(z.matrix)))
```

#### MASE
```{r}
dplyr::filter(eval.df, `LVs used in reconstruction` == 
                paste("All, n =", ncol(z.matrix))) %>%
  ggplot2::ggplot(ggplot2::aes(x = MASE,
                               group = Dataset, fill = Dataset)) +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::theme_bw() +
  ggplot2::scale_fill_manual(values = c("white", "gray50", "black")) +
  ggplot2::ggtitle(paste("All, n =", ncol(z.matrix)))
```

### Summary

The isolated immune cell samples (`SRP045500`) are among some of the best 
reconstructed (high correlation, low error) in the recount2 compendium.
`E-MTAB-2452`, the microarray dataset, is reconstructed relatively poorly. 
This is unsurprising, as it was not included in training and is from a different
technology.
We'll look next at reconstruction using _subsets_ of the recount2 PLIER model
LVs. 
Specifically, we hypothesize that reconstruction using only pathway-associated
LVs (e.g., those LVs rooted in biological signal) may make the measures of
reconstruction performance -- correlation and error -- more similar between
the two isolated immune cell datasets.
If this is the case, it supports the idea that the PLIER-learned LVs that are 
not associated with pathways are, at least in part, capturing variation that
can be attributed to technical factors.
We would expect that the technical signals in recount2 would be distinct 
(e.g., an LV associated with library prep strategy) from what's in 
`E-MTAB-2452` and, therefore, that including this information may negatively 
impact reconstruction of this dataset.

## Reconstruction of E-MTAB-2452 with pathway-associated LVs only

```{r}
# Get the significant LVs (from recount2 model) using 0.05 as the FDR cutoff
plier.summary <- plier.results$summary
sig.summary <- plier.summary %>%
                  dplyr::filter(FDR < 0.05)
sig.lvs <- unique(sig.summary$`LV index`)
```

```{r}
# drop columns (LVs) from Z that are not significantly associated with prior 
# info
sig.z.mat <- z.matrix[, as.integer(sig.lvs)]
# drop rows (LVs) from E-MTAB-2452 B that are not significantly associated with 
# prior info
iso.sig.b.mat <- as.matrix(iso.b.matrix[as.integer(sig.lvs), ])
```

### Reconstruction and evaluation

```{r}
path.lv.type <- paste("Pathway-associated, n =", ncol(sig.z.mat))
iso.recon.sig.df <- ReconstructionEvalWrapper(z.mat = as.matrix(sig.z.mat),
                                              b.mat = iso.sig.b.mat,
                                              input.exprs = iso.ord.rownorm,
                                              dataset.name = "E-MTAB-2452",
                                              lv.type = path.lv.type)
head(iso.recon.sig.df)
```
Add to rest of evaluations

```{r}
eval.df <- dplyr::bind_rows(eval.df, iso.recon.sig.df)
rm(iso.recon.sig.df)
```

## Reconstruction with LVs that are not associated with pathways

Here, we'll have to reconstruct the recount2 data with the LVs that are not
significantly associated with pathways, as we did not perform this analysis
in `02-recount2_PLIER_exploration`.

```{r}
# some LVs don't even make it to the summary data.frame from the PLIER model
# the LVs we want here are *all* the LVs that do not have at least one pathway 
# significantly associated with them
all(1:nrow(plier.results$B) %in% unique(plier.summary$`LV index`))
```

```{r}
# non-significant LVs, then, are all LVs that are not in the list of 
# pathway-associated LVs
non.lvs <- setdiff(1:nrow(plier.results$B), as.integer(sig.lvs))
```
```{r}
# z matrix for just these LVs
non.z.mat <- z.matrix[, as.integer(non.lvs)]
```
### recount2

```{r}
# get B matrix
non.b.mat <- as.matrix(plier.results$B[as.integer(non.lvs), ])
# reconstruction & evaluation 
non.lv.type <- paste("LVs not associated with pathway, n =", length(non.lvs))
recount.recon.non.df <- 
  ReconstructionEvalWrapper(z.mat = as.matrix(non.z.mat),
                            b.mat = non.b.mat,
                            input.exprs = recount.input.exprs,
                            lv.type = non.lv.type,
                            dataset.name = "recount2")
head(recount.recon.non.df)
```

#### SRP045500

```{r}
srp.non.df <- recount.recon.non.df %>%
            dplyr::filter(grepl("SRP045500", recount.recon.non.df$Sample)) %>%
            dplyr::mutate(Dataset = "SRP045500")
recount.recon.non.df <- dplyr::bind_rows(recount.recon.non.df, srp.non.df)
```

```{r}
eval.df <- dplyr::bind_rows(eval.df, recount.recon.non.df)
rm(recount.recon.non.df, srp.non.df)
```

### E-MTAB-2452

```{r}
# get b matrix
iso.non.b.mat <- as.matrix(iso.b.matrix[as.integer(non.lvs), ])
# reconstruction & evaluation 
iso.recon.non.df <- 
  ReconstructionEvalWrapper(z.mat = as.matrix(non.z.mat),
                            b.mat = iso.non.b.mat,
                            input.exprs = iso.ord.rownorm,
                            lv.type = non.lv.type,
                            dataset.name = "E-MTAB-2452")
head(iso.recon.non.df)
```

```{r}
eval.df <- dplyr::bind_rows(eval.df, iso.recon.non.df)
rm(iso.recon.non.df)
```

## Plotting

Plot all three conditions on the same plot for easy comparison

### Density

#### Correlation 
```{r}
ggplot2::ggplot(eval.df,
  ggplot2::aes(x = `Spearman correlation`, group = Dataset, fill = Dataset)) +
  ggplot2::facet_wrap(~ `LVs used in reconstruction`, ncol = 1) +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::theme_bw() +
  ggplot2::scale_fill_manual(values = c("white", "gray50", "black"))
```

```{r}
plot.file <- file.path(plot.dir, 
                       "recount2_model_isolated_cell_recon_correlation.pdf")
ggplot2::ggsave(filename = plot.file, plot = ggplot2::last_plot(),
                height = 11, width = 8.5)
```

#### Error
```{r}
ggplot2::ggplot(eval.df,
  ggplot2::aes(x = MASE, group = Dataset, fill = Dataset)) +
  ggplot2::facet_wrap(~ `LVs used in reconstruction`, ncol = 1) +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::theme_bw() +
  ggplot2::scale_fill_manual(values = c("white", "gray50", "black"))
```

```{r}
plot.file <- file.path(plot.dir, 
                       "recount2_model_isolated_cell_recon_error.pdf")
ggplot2::ggsave(filename = plot.file, plot = ggplot2::last_plot(),
                height = 11, width = 8.5)
```

### E-MTAB-2452 Boxplots

#### Correlation
```{r}
dplyr::filter(eval.df, Dataset == "E-MTAB-2452") %>%
  ggplot2::ggplot(ggplot2::aes(x = `LVs used in reconstruction`,
                               y = `Spearman correlation`)) +
  ggplot2::geom_boxplot() +
  ggplot2::theme_bw() +
  
  ggplot2::ggtitle("E-MTAB-2452 reconstruction with recount2 model")
```
```{r}
plot.file <- file.path(plot.dir,
                       "E-MTAB-2452_reconstruction_corr_recount2_model.pdf")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot())
```
#### Error
```{r}
dplyr::filter(eval.df, Dataset == "E-MTAB-2452") %>%
  ggplot2::ggplot(ggplot2::aes(x = `LVs used in reconstruction`,
                               y = MASE)) +
  ggplot2::geom_boxplot() +
  ggplot2::theme_bw() +
  ggplot2::scale_x_discrete(labels = c("All", "Not pathway-associated", 
                                       "Pathway-associated")) +
  ggplot2::ggtitle("E-MTAB-2452 reconstruction with recount2 model")
```
```{r}
plot.file <- file.path(plot.dir,
                       "E-MTAB-2452_reconstruction_error_recount2_model.pdf")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot())
```

## Summary

Reconstruction of a test dataset, `E-MTAB-2452`, using only PLIER latent 
variables that significantly associated with a pathway or gene set is 
better in terms of both MASE and Spearman correlation than when using only
latent variables that are not associated with any prior information.
This is despite the fact that there are more non-significant LVs (`n = 788` vs.
`n = 199`).
However, neither of these subsets of LVs outperforms using all LVs
and the performance gap between reconstructing `E-MTAB-2452` and `SRP045500` 
(the sorted immune cell dataset included in recount2) is not considerably 
reduced when using only pathway-associated LVs, which refutes our initial
hypothesis.
It's also interesting that, when reconstructing using only non-significant LVs, 
the pre- and post-reconstruction correlation values are much more similar 
between `E-MTAB-2452` and `SRP045500`.
This suggests that the pathway-associated LVs are important for accurate
reconstruction of the `SRP045500` dataset.

_It's worth noting that these sorted immune cell datasets are from 
patients with different diagnoses and contain different cell type populations 
(in addition to different technologies) that may be more or less 
well-represented in the recount2 compendium.
We expect that these differences contribute to the results. 
Future analyses may focus on a microarray dataset more comparable to `SRP045500`
in these ways._
