---
title: "Systemic lupus erythematosus whole blood compendium PLIER training"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

In this notebook, we'll train a PLIER model on the systemic lupus erythematosis 
(SLE) whole blood (WB) compendium we processed in 
[`greenelab/rheum-plier-data/sle-wb`](https://github.com/greenelab/rheum-plier-data/tree/4be547553f24fecac9e2f5c2b469a17f9df253f0/sle-wb)
and do some exploratory data analysis.

## Functions and directory set up

```{r}
library(AnnotationDbi)

`%>%` <- dplyr::`%>%`
# custom functions
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "05")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "05")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Load SLE WB data
```{r}
exprs.file <- file.path("data", "expression_data", 
                        "SLE_WB_all_microarray_QN_zto_before.pcl")
exprs.df <- readr::read_tsv(exprs.file)
```
### Annotation
```{r}
symbol.obj <- org.Hs.eg.db::org.Hs.egSYMBOL
mapped.genes <- AnnotationDbi::mappedkeys(symbol.obj)
symbol.list <- as.list(symbol.obj[mapped.genes])
symbol.df <- as.data.frame(cbind(names(symbol.list), unlist(symbol.list)))
colnames(symbol.df) <- c("EntrezID", "GeneSymbol")

# get gene column name to match to facilitate use with dplyr
colnames(exprs.df)[1] <- "EntrezID"

# matching types
symbol.df$EntrezID <- as.integer(as.character(symbol.df$EntrezID))

# inner join
annot.exprs.df <- dplyr::inner_join(symbol.df, exprs.df, by = "EntrezID")

symbol.file <- 
  file.path("data", "expression_data", 
            "SLE_WB_all_microarray_QN_zto_before_with_GeneSymbol.pcl")

readr::write_delim(annot.exprs.df, path = symbol.file, delim = "\t")

# matrix with gene symbol as rownames
exprs.mat <- dplyr::select(annot.exprs.df, -EntrezID)
rownames(exprs.mat) <- exprs.mat$GeneSymbol
exprs.mat <- as.matrix(dplyr::select(exprs.mat, -GeneSymbol))
```
```{r}
exprs.mat[1:5, 1:5]
```

## PLIER model training

```{r}
plier.result <- PLIERNewData(exprs.mat = exprs.mat)
```
```{r}
model.file <- file.path(results.dir, "SLE-WB_PLIER_model.RDS")
saveRDS(plier.result, file = model.file)
```

## Explore SLE WB PLIER model

### U matrix

We can get an overview of what pathways are captured with the LVs by plotting 
`U` with `PLIER::plotU`.

```{r}
pdf(file.path(plot.dir, "SLE-WB_PLIER_Uplot_auc0.75.pdf"))
PLIER::plotU(plier.result, auc.cutoff = 0.75, fontsize_row = 4,
             fontsize_col = 7)
dev.off()
```
#### All LVs

What proportion of entries in `U` are non-zero (column-wise)?
As a reminder, the fewer pathways that are associated with an LV (column), the
easier interpretation will be. 
(Sparsity of `U` is a constraint in PLIER.)

```{r}
num.lvs <- ncol(plier.result$Z)
u.sparsity.all <- CalculateUSparsity(plier.results = plier.result,
                                     significant.only = FALSE)

ggplot2::ggplot(as.data.frame(u.sparsity.all),
                ggplot2::aes(x = u.sparsity.all)) +
  ggplot2::geom_density(fill = "blue", alpha = 0.5) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "proportion of positive entries in U",
                title = paste("All LVs, n =", num.lvs))
```

```{r}
plot.file <- file.path(plot.dir,
                       "SLE-WB_model_U_sparsity_all_density.png")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot(),
                height = 5, width = 7)
```
```{r}
summary(u.sparsity.all)
```

#### Pathway-associated LVs

Now, only with _significant_ associations (`FDR < 0.05`).

```{r}
plier.summary <- plier.result$summary
sig.summary <- plier.summary %>%
  dplyr::filter(FDR < 0.05)
num.sig.lvs <- length(unique(sig.summary$`LV index`))
```

```{r}
u.sparsity.sig <- CalculateUSparsity(plier.results = plier.result,
                                     significant.only = TRUE,
                                     fdr.cutoff = 0.05)

ggplot2::ggplot(as.data.frame(u.sparsity.sig),
                ggplot2::aes(x = u.sparsity.sig)) +
  ggplot2::geom_density(fill = "blue", alpha = 0.5) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "proportion of positive entries in U",
                title = paste("Pathway-associated LVs, n =", num.sig.lvs))
```
```{r}
plot.file <- file.path(plot.dir,
                       "SLE-WB_model_U_sparsity_sig_density.png")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot(),
                height = 5, width = 7)
```

```{r}
summary(u.sparsity.sig)
```

### Pathway coverage

What proportion of the pathways input into the model are significantly 
associated (`FDR < 0.05`) with an LV? 

```{r}
sle.coverage <- GetPathwayCoverage(plier.results = plier.result)

# Pathway coverage
sle.coverage$pathway
```

Less pathways are "covered" with this model than the recount2 model, where 
coverage was `0.419`.

### PCA of lower-dimensional (LV) space

We'll perform PCA on the B matrix from the SLE WB PLIER model. 
This is mainly for visualization purposes. 

As noted when we prepped the data in [`greenelab/rheum-plier-data/sle-wb`](https://github.com/greenelab/rheum-plier-data/tree/4be547553f24fecac9e2f5c2b469a17f9df253f0/sle-wb), [there are still differences](https://github.com/greenelab/rheum-plier-data/blob/4be547553f24fecac9e2f5c2b469a17f9df253f0/sle-wb/plots/PCA/SLE_WB_all_microarray_QN_PC1-5_zto.before.png) between datasets in the final compendium.

If the LVs that are not associated with any pathway (all `AUC < 0.75`) 
in fact capture nuisance variables, we expect that dropping them from 
the `B` matrix may reduce the dataset effect.

```{r}
# Which LVs have pathways with AUC > 0.75?
lv.auc <- 
  as.integer(unique(plier.summary$`LV index`[which(plier.summary$AUC > 0.75)]))

# B matrix
b.matrix <- as.matrix(plier.result$B)

# B matrix -- only LV AUC > 0.75 (any pathway)
sig.b.mat <- b.matrix[lv.auc, ]

# B matrix all other LV 
oth.b.mat <- b.matrix[-lv.auc, ]
```

```{r}
# color palette for PCA plots
plot.color.pal <- c("#54FF9F", "#43CD80", "#2E8B57", "#006400", "#FF8C00",
                    "#8B4500", "#000080")
# read sample dataset mapping file
sd.file <- file.path("data", "sample_info", "sle-wb_sample_dataset_mapping.tsv")
sd.df <- readr::read_tsv(sd.file)
```

#### All latent variables
```{r}
# PCA
all.lv.pc <- prcomp(t(b.matrix))
cum.var.exp <- cumsum(all.lv.pc$sdev^2 / sum(all.lv.pc$sdev^2))

# PC1-2 in form suitable for ggplot2
all.lv.df <- as.data.frame(cbind(rownames(all.lv.pc$x),
                                 all.lv.pc$x[, 1:2]))
colnames(all.lv.df)[1] <- "SampleID"
# add dataset of origin info
all.lv.df <- dplyr::full_join(all.lv.df, sd.df, by = "SampleID") %>%
  dplyr::mutate(PC1 = as.numeric(as.character(PC1)),
                PC2 = as.numeric(as.character(PC2)),
                Dataset = factor(Dataset, 
                                 levels = c("E-GEOD-39088",
                                            "E-GEOD-61635",
                                            "E-GEOD-72747",
                                            "E-GEOD-11907",
                                            "E-GEOD-49454",
                                            "E-GEOD-65391",
                                            "E-GEOD-78193")))

# plotting
all.plot <- 
  ggplot2::ggplot(all.lv.df, ggplot2::aes(x = PC1, y = PC2, colour = Dataset)) +
    ggplot2::geom_point(alpha = 0.5) +
    ggplot2::scale_color_manual(values = plot.color.pal) +
    ggplot2::theme_bw() +
    ggplot2::labs(x = paste0("PC1 (cum var exp = ", 
                             round(cum.var.exp[1], 3), ")"),
                  y = paste0("PC2 (cum var exp = ", 
                             round(cum.var.exp[2], 3), ")"), 
                  title = "All Latent Variables") +
    ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, 
                                                      face = "bold"))
all.plot
```

#### Pathway-associated latent variables
```{r}
# PCA
path.lv.pc <- prcomp(t(sig.b.mat))
cum.var.exp <- cumsum(path.lv.pc$sdev^2 / sum(path.lv.pc$sdev^2))
# PC1-2 in form suitable for ggplot2
path.lv.df <- as.data.frame(cbind(rownames(path.lv.pc$x),
                                 path.lv.pc$x[, 1:2]))
colnames(path.lv.df)[1] <- "SampleID"
# add dataset of origin info
path.lv.df <- dplyr::full_join(path.lv.df, sd.df, by = "SampleID") %>%
  dplyr::mutate(PC1 = as.numeric(as.character(PC1)),
                PC2 = as.numeric(as.character(PC2)),
                Dataset = factor(Dataset, 
                                 levels = c("E-GEOD-39088",
                                            "E-GEOD-61635",
                                            "E-GEOD-72747",
                                            "E-GEOD-11907",
                                            "E-GEOD-49454",
                                            "E-GEOD-65391",
                                            "E-GEOD-78193")))

# plotting
path.plot <- 
  ggplot2::ggplot(path.lv.df, ggplot2::aes(x = PC1, y = PC2, 
                                           colour = Dataset)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::scale_color_manual(values = plot.color.pal) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = paste0("PC1 (cum var exp = ", 
                           round(cum.var.exp[1], 3), ")"),
                y = paste0("PC2 (cum var exp = ", 
                           round(cum.var.exp[2], 3), ")"),
                title = "Pathway-associated Latent Variables (AUC > 0.75)") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, 
                                                    face = "bold"))
path.plot
```

#### Other latent variables (AUC <= 0.75)
```{r}
# PCA
oth.lv.pc <- prcomp(t(oth.b.mat))
cum.var.exp <- cumsum(oth.lv.pc$sdev^2 / sum(oth.lv.pc$sdev^2))
# PC1-2 in form suitable for ggplot2
oth.lv.df <- as.data.frame(cbind(rownames(oth.lv.pc$x),
                                 oth.lv.pc$x[, 1:2]))
colnames(oth.lv.df)[1] <- "SampleID"
# add dataset of origin info
oth.lv.df <- dplyr::full_join(oth.lv.df, sd.df, by = "SampleID") %>%
  dplyr::mutate(PC1 = as.numeric(as.character(PC1)),
                PC2 = as.numeric(as.character(PC2)),
                Dataset = factor(Dataset, 
                                 levels = c("E-GEOD-39088",
                                            "E-GEOD-61635",
                                            "E-GEOD-72747",
                                            "E-GEOD-11907",
                                            "E-GEOD-49454",
                                            "E-GEOD-65391",
                                            "E-GEOD-78193")))

# plotting
oth.plot <- 
  ggplot2::ggplot(oth.lv.df, ggplot2::aes(x = PC1, y = PC2, colour = Dataset)) +
    ggplot2::geom_point(alpha = 0.5) +
    ggplot2::scale_color_manual(values = plot.color.pal) +
    ggplot2::theme_bw() +
    ggplot2::labs(x = paste0("PC1 (cum var exp = ", 
                           round(cum.var.exp[1], 3), ")"),
                  y = paste0("PC2 (cum var exp = ", 
                           round(cum.var.exp[2], 3), ")"),
                  title = "Latent Variables AUC <= 0.75") +
    ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, 
                                                      face = "bold"))
oth.plot
```

```{r}
# Save to PDF
pdf(file.path(plot.dir, "SLE-WB_PLIER_LV_PCA_plots.pdf"), height = 14, 
              width = 7)
gridExtra::grid.arrange(all.plot, path.plot, oth.plot, ncol = 1)
dev.off()
```