---
title: "SLE WB PLIER cell type analyses"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---
**J. Taroni 2018**

Part of the motivation behind using PLIER (particularly in a systemic lupus 
erythematosus whole blood setting) is that we can automatically extract 
information about cell type proportion variation. 
One of the datasets included in the SLE WB compendium comes from 
[Banchereau, et al.](https://doi.org/10.1016/j.cell.2016.03.008).
The original publication contained some interesting findings about neutrophils 
and plasmablasts.
In this notebook, we explore whether latent variables from the PLIER model 
trained on the SLE WB compendium associated with these cell type (gene sets) 
show similar trends to what was found by Banchereau, et al.
By extension, this also tells us whether the cell type-specific patterns are 
retained following all the processing we did in 
[`greenelab/rheum-plier-data/sle-wb`](https://github.com/greenelab/rheum-plier-data/tree/4be547553f24fecac9e2f5c2b469a17f9df253f0/sle-wb).

## Functions and directory set up

```{r}
`%>%` <- dplyr::`%>%`
library(PLIER)

# custom functions
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "06")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "06")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```


## Load data

### PLIER model
```{r}
# SLE WB model trained in notebook 05
model.file <- file.path("results", "05", "SLE-WB_PLIER_model.RDS")
plier.results <- readRDS(model.file)
# summary data.frame (pathway to LV info like AUC, FDR)
sle.summary <- plier.results$summary
# LVs
b.matrix <- plier.results$B
```

### Banchereau, et al. Sample Data Relationship File
```{r}
# phenotype information 
# read in sample data relationship file
file.65391 <- file.path("data", "sample_info", "E-GEOD-65391.sdrf.txt")
e.65391.sdrf <- readr::read_tsv(file.65391)
# use Sample instead of Source Name
colnames(e.65391.sdrf)[1] <- c("Sample")
# get rid of trailing " 1" so they match the sample names from B matrix
e.65391.sdrf$Sample <- gsub(" 1", "", e.65391.sdrf$Sample)
```

## Neutrophil signature

### LVs associated with neutrophil gene sets

```{r}
sle.summary %>%
  dplyr::filter(grepl("Neutrophil", pathway), 
                FDR < 0.05)
```
Based on these results, we select `LV2`,`LV27`, `LV34`, and `LV87` 
for further exploration.
If these latent variables are also strongly associated with _other_ pathways or 
cell types, that is not desirable. 
That could make intepretation more difficult.

```{r}
sle.summary %>%
  dplyr::filter(`LV index` == 2 |
                  `LV index` == 27 |
                  `LV index` == 34 |
                  `LV index` == 87) %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::arrange(`LV index`, desc(AUC))
```

Some notes on these results: 

* `LV2` looks like it captures the myeloid lineage in a broad sense, as a
monocyte gene set, `SVM Monocytes`, and a neutrophil progenitor gene set,
`DMAP_GRAN2` (reported to be a neutrophilic metamyelocyte signature in
the [DMAP paper](https://doi.org/10.1016/j.cell.2011.01.004)).

* `LV27` and `LV34` also look like they may capture some information about
monocytes because of their association with the `DMAP_MONO2` gene set.

* `LV87` looks the most "neutrophil-specific" and therefore, the most desirable
for this kind of analysis.

#### U matrix

```{r}
PLIER::plotU(plierRes = plier.results,
             pval.cutoff = 1e-06,
             indexCol = c(2, 27, 34, 87),
             top = 10)
```
```{r}
png(file.path(plot.dir, "SLE_WB_neutrophil_Uplot.png"), 
    res = 300, width = 7, height = 7, units = "in")
PLIER::plotU(plierRes = plier.results,
             pval.cutoff = 1e-06,
             indexCol = c(2, 27, 34, 87),
             top = 10)
dev.off()
```

### Compare to neutrophil count from Banchereau, et al. 

```{r}
neutrophil.count.df <- e.65391.sdrf[, c("Sample", 
                                        "Characteristics [neutrophil_count]")]
colnames(neutrophil.count.df)[2] <- c("Neutrophil.Count")
neutrophil.count.df <- 
  neutrophil.count.df %>%
  dplyr::filter(Neutrophil.Count != "Data Not Available") %>%
  dplyr::filter(Neutrophil.Count != "Not Applicable")

```


```{r}
# combine with neutrophil latent variables
neutro.lv.df <- as.data.frame(cbind(colnames(b.matrix), 
                                    t(b.matrix[c(2, 27, 34, 87), ])))
colnames(neutro.lv.df) <- c("Sample", "LV2", "LV27", "LV34", "LV87")

neutro.df <- dplyr::inner_join(neutro.lv.df, neutrophil.count.df,
                               by = "Sample") %>%
  dplyr::mutate(LV2 = as.numeric(as.character(LV2)),
                LV27 = as.numeric(as.character(LV27)),
                LV34 = as.numeric(as.character(LV34)),
                LV87 = as.numeric(as.character(LV87)),
                Neutrophil.Count = as.numeric(as.character(Neutrophil.Count)))
neutro.df
```

```{r}
# write to file, will use to compare to results with recount2
count.file <- file.path(results.dir, 
                        "Banchereau_count_SLE_model_neutrophil_LV.tsv")
readr::write_tsv(neutro.df, path = count.file)
```


```{r}
# function for scatter plots, given lv (a string indicating the LV of interest)
# make a scatter plot where the LV is the x variable, neutrophil count is the
# y variable & fit a line with geom_smooth(method = "lm")
# needs to be used in this global environment
LVScatter <- function(lv = "LV2") {
  y.var <- "Neutrophil.Count"
  ggplot2::ggplot(neutro.df, ggplot2::aes_string(x = lv, y = y.var)) +
  ggplot2::geom_count() +
  ggplot2::geom_smooth(method = "lm") +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "Neutrophil Count") +
  ggplot2::theme(legend.position = "none")
}
```

#### LV2

```{r}
LVScatter(lv = "LV2")
```

```{r}
summary(lm(neutro.df$Neutrophil.Count ~ neutro.df$LV2))
```

#### LV27

```{r}
LVScatter(lv = "LV27")
```

```{r}
summary(lm(neutro.df$Neutrophil.Count ~ neutro.df$LV27))
```

#### LV34

```{r}
LVScatter(lv = "LV34")
```

```{r}
summary(lm(neutro.df$Neutrophil.Count ~ neutro.df$LV34))
```

#### LV87

```{r}
LVScatter(lv = "LV87")
```

```{r}
summary(lm(neutro.df$Neutrophil.Count ~ neutro.df$LV87))
```

When comparing the results to the recount2 model, we'll pursue `LV27` because 
it's the best performing and `LV87` because of its lack of overlap with
monocyte signatures.
So, we'll save the plots for these two LVs.

```{r}
plot.file <- file.path(plot.dir, "Banchereau_neutrophil_count_LV27_scatter.png")
ggplot2::ggsave(plot.file, plot = LVScatter("LV27"))
plot.file <- file.path(plot.dir, "Banchereau_neutrophil_count_LV87_scatter.png")
ggplot2::ggsave(plot.file, plot = LVScatter("LV87"))
```

## Plasma cell signature

In Banchereau, et al., the authors demonstrated that plasmablast counts were 
different between patients stratified by 3 disease activity (DA) groups. 

While there were no _plasmablast_ gene sets given to PLIER during training, we 
did use _plasma cell_ gene sets (isolated from PBMCs)
(A plasmablast is a plasma cell precursor that is typically found in a
germinal center.)

### LVs associated with plasma cell gene sets

What LVs, if any, are significantly associated with the 

```{r}
sle.summary %>%
  dplyr::filter(grepl("Plasma", pathway), 
                FDR < 0.05)
```
```{r}
sle.summary %>% 
  dplyr::filter(`LV index` == 52 |
                 `LV index` == 136,
                FDR < 0.05) %>%
  dplyr::arrange(desc(AUC))
```

#### U matrix
```{r}
PLIER::plotU(plierRes = plier.results,
             pval.cutoff = 1e-06,
             indexCol = c(52, 136),
             top = 10)
```

```{r}
png(file.path(plot.dir, "SLE_WB_plasma_cell_Uplot.png"), 
    res = 300, width = 7, height = 7, units = "in")
PLIER::plotU(plierRes = plier.results,
             pval.cutoff = 1e-06,
             indexCol = c(52, 136),
             top = 10)
dev.off()
```


### Banchereau, et al. disease activity groups

```{r}
# get DA information from the sample-data relationship file
da.group.df <- e.65391.sdrf[, c("Sample", "Characteristics [disease_activity]")]
colnames(da.group.df)[2] <- "Disease.Activity"
da.group.df <- da.group.df %>%
              dplyr::filter(Disease.Activity != "Not Applicable")

# plasma cell LV
plasma.lv.df <- as.data.frame(cbind(colnames(b.matrix), 
                                    t(b.matrix[c(52, 136), ])))
colnames(plasma.lv.df) <- c("Sample", "LV52", "LV136")

# join
plasma.df <- dplyr::inner_join(plasma.lv.df, da.group.df) %>%
  dplyr::mutate(Disease.Activity = dplyr::recode(Disease.Activity, 
                                                 `1` = "DA1", 
                                                 `2` = "DA2",
                                                 `3` = "DA3"),
                Disease.Activity = factor(Disease.Activity),
                LV52 = as.numeric(as.character(LV52)),
                LV136 = as.numeric(as.character(LV136))) 
head(plasma.df)
```

```{r}
# write to file, will be used to compare to recount2 model
plasma.file <- file.path(results.dir, 
                         "Banchereau_DA_group_SLE_model_plasma_cell_LVs.tsv")
readr::write_tsv(plasma.df, path = plasma.file)
```

#### Plotting
```{r}
plasma.df %>%
  ggplot2::ggplot(ggplot2::aes(x = Disease.Activity, 
                               y = LV52)) +
  ggplot2::geom_boxplot(notch = TRUE) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "Disease Activity")
```

```{r}
plot.file <- file.path(plot.dir, "Banchereau_DA_group_LV52_boxplot.png")
ggplot2::ggsave(filename = plot.file, plot = ggplot2::last_plot())
```

```{r}
# check for statistical significance 
# (pairwise t-test is consistent with original publication as far as I can tell)
pairwise.t.test(x = plasma.df$LV52, 
                g = plasma.df$Disease.Activity, 
                p.adjust.method = "bonferroni")
```

```{r}
plasma.df %>%
  ggplot2::ggplot(ggplot2::aes(x = Disease.Activity, 
                               y = LV136)) +
  ggplot2::geom_boxplot(notch = TRUE) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "Disease Activity")
```

```{r}
plot.file <- file.path(plot.dir, "Banchereau_DA_group_LV136_boxplot.png")
ggplot2::ggsave(filename = plot.file, plot = ggplot2::last_plot())
```

```{r}
# check for statistical significance 
# (pairwise t-test is consistent with original publication as far as I can tell)
pairwise.t.test(x = plasma.df$LV136, 
                g = plasma.df$Disease.Activity, 
                p.adjust.method = "bonferroni")
```