---
title: "SLE IFN modulatory therapies: Analysis"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

In this notebook, we'll examine how IFN-inducible or IFN-associated gene 
signatures change during treatment with targeted therapies that are designed to 
block the action of some IFNs.

Briefly, we'll look at two therapies in the context of SLE: IFN-K (which blocks 
IFN-alpha, type I IFN; 
[Lauwerys, et al. _Arthritis Rheum._ 2013.](https://doi.org/10.1002/art.37785)) 
and AMG 811 (which blocks IFN-gamma, type II IFN; 
[Welcher, et al. _Arthritis Rheumatol._ 2015.](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5054935/)).

For a little more background, including summaries of the main findings in the 
original papers, see the `7-sle_ifn_data_prep` notebook.

## Directory setup

```{r}
`%>%` <- dplyr::`%>%`
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "10")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "10")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
# directory that was used for data prep
data.prep.dir <- file.path("results", "09")
```

```{r}
# set seed for reproducibility (plot jitter)
set.seed(123)
```

## IFN-K treatment

**E-GEOD-39088; Lauwerys, et al. 2013.**

### Functions

In the IFN-K trial specifically, we'll examine how IFN expression has 
changed from baseline.
This is similar to the analyses performed in the original paper. 
We need a custom function for this and for plotting, as we'll be repeating this
with three different methods: modular framework, SLE WB PLIER, recount2 PLIER.

```{r}
CalculateChangeFromBaseline <- function(df, variable.id, value.id) {
  # Given a data.frame that contains values (value.id) from the longitudinal 
  # study E-GEOD-39088, this function will calculate the change in expression
  # level from baseline. Not intended for use outside this context (& env)!
  #
  # Args:
  #   df: data.frame that contains day, some variable (i.e., module or latent
  #       variable id), and the value for that variable (i.e., mean expression);
  #       must contain Day and Patient columns
  #   variable.id: the name of the column that contains the module or latent
  #               variable identifier 
  #   value.id: the name of the column that contains the values to be subtracted
  #             from one another (e.g., contains mean expression for a sample
  #             for above variable)
  # 
  # Returns:
  #  df with the calculated changes appended in the "Change" column
  
  col.check <- all(("Day" %in% colnames(df)), ("Patient" %in% colnames(df)))
  if (!col.check) {
    stop("This function expects the input data.frame to contain colnames 
         'Day' and 'Patient'")
  }
  
  input.col.check <- all((variable.id %in% colnames(df)), 
                          (value.id %in% colnames(df)))
  if (!input.col.check) {
    stop("One or more of 'variable.id' and 'value.id' are not in
         colnames(df)")
  }
  
  if (!("baseline" %in% df$Day)) {
    stop("baseline should be a time point in Day column")
  }
  
  change.summary <- rep(NA, nrow(df))
  # for each patient, how has the value changed?
  for (pat in unique(df$Patient)) {
    # for each variable (e.g., module)
    for(var.iter in unique(df[, variable.id])) {
      # identify the baseline value index
      baseline.indx <- which(df[, variable.id] == var.iter & 
                               df$Patient == pat &
                               df$Day == "baseline")
      # for all days, including baseline (baseline should equal zero)
      for(day in unique(df$Day)) {
        day.indx <- which(df[, variable.id] == var.iter & 
                            df$Patient == pat &
                            df$Day == day)
        # subtract the baseline value from the day value
        change.summary[day.indx] <- 
          df[, value.id][day.indx] - df[, value.id][baseline.indx]
      }
    }
  }
  
  # add this column to the data.frame
  df$Change <- change.summary 
  return(df)
}

PlotChangeFromBaseline <- function(df, y.label, plot.title, plot.path,
                                   facets = "Module ~ Day",
                                   plot.subtitle = "Lauwerys, et al.") {
  # Given a data.frame that with calculated changes from baseline 
  # (from CalculateChangeFromBaseline), make boxplots comparing the three groups
  # of patients -- placebo, IFN-negative, IFN-positive -- for each variable
  # (e.g. module) ~ each day of the study (e.g., day 112, day 168)
  # Not intended for use outside this context (& env)!
  #
  # Args:
  #   df: data.frame output from CalculateChangeFromBaseline
  #   y.label: y-axis label (string)
  #   plot.title: plot title (string)
  #   plot.path: full path for plot file (string)
  #   facets: formula passed to ggplot2::facet_wrap(), default is "Module ~ Day"
  #   plot.subtitle: plot subtitle (string)
  #
  # Returns:
  #   NULL; plot saved at plot.path
  
  # reorder IFN-level for display
  df$`IFN-level` <- 
   factor(df$`IFN-level`,
          levels = c("Placebo", "IFN-negative", "IFN-positive"))

  # plot
  p <- ggplot2::ggplot(dplyr::filter(df, Day != "baseline"), 
                       ggplot2::aes(x = `IFN-level`, 
                                    y = `Change`, color = `IFN-level`)) +
    ggplot2::geom_boxplot(notch = TRUE) + 
    ggplot2::theme_bw() + 
    ggplot2::geom_jitter(alpha = 0.5, width = 0.2) +
    ggplot2::facet_wrap(as.formula(facets), ncol = 2) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                   legend.position = "none", 
                   text = ggplot2::element_text(size = 15)) +
    ggplot2::labs(y = y.label,
                  title = plot.title,
                  subtitle = plot.subtitle) +
    ggplot2::scale_color_manual(values = c("#969696", "#8073ac", "#e08214"))
  
  ggplot2::ggsave(plot.path, plot = p, width = 8.5, height = 14, units = "in")
}
```

### Modular transcriptional analyses

```{r}
# read in tidy data
mod.file <- file.path(data.prep.dir, "E-GEOD-39088_Chiche_et_al_module.tsv")
mod.summary.df <- readr::read_tsv(mod.file)
```

```{r}
# only keep SLE patients
ifn.summary.df <-   
  mod.summary.df %>% 
    dplyr::filter(grepl("SLE patient", Patient))

ifn.summary.df <- 
  CalculateChangeFromBaseline(df = as.data.frame(ifn.summary.df),
                              variable.id = "Module",
                              value.id = "Summary")
df.file <- file.path(results.dir, "E-GEOD-39088_IFNk_Chiche_modules_change.tsv")
readr::write_tsv(ifn.summary.df, df.file)
```

#### Plot

```{r}
plot.file <- file.path(plot.dir, "E-GEOD-39088_IFNk_Chiche_modules_change.pdf")
PlotChangeFromBaseline(df = ifn.summary.df,
                       y.label = "Change in Expression Summary",
                       plot.title = "IFN Modular Framework Expression - 
                          IFN-K treatment",
                       plot.path = plot.file)
```

```{r}
rm(list = setdiff(ls(), c("%>%", "CalculateChangeFromBaseline",
                          "PlotChangeFromBaseline",
                          "plot.dir", "results.dir", "data.prep.dir")))
```

### PLIER trained on SLE WB compendium

```{r}
sle.b.file <- file.path(data.prep.dir, "E-GEOD-39088_SLE-WB_PLIER_IFN_B.tsv")
sle.b.df <- readr::read_tsv(sle.b.file)
```

```{r}
# SLE patients only
ifn.b.df <- sle.b.df %>%
  dplyr::filter(grepl("SLE patient", Patient))

ifn.b.df <- CalculateChangeFromBaseline(df = as.data.frame(ifn.b.df),
                                        variable.id = "LV",
                                        value.id = "Value")
df.file <- file.path(results.dir, "E-GEOD-39088_IFNk_SLE_PLIER_change.tsv")
readr::write_tsv(ifn.b.df, df.file)
```

```{r}
plot.file <- file.path(plot.dir, "E-GEOD-39088_IFNk_SLE_PLIER_change.pdf")
PlotChangeFromBaseline(df = ifn.b.df,
                       y.label = "Change in LV value",
                       plot.title = "SLE WB PLIER - IFN-K treatment",
                       plot.path = plot.file,
                       facets = "LV ~ Day")
```

```{r}
rm(list = setdiff(ls(), c("%>%", "CalculateChangeFromBaseline",
                          "PlotChangeFromBaseline",
                          "plot.dir", "results.dir", "data.prep.dir")))
```

### PLIER trained on `recount2`

```{r}
recount.b.file <- file.path(data.prep.dir, 
                            "E-GEOD-39088_recount2_PLIER_IFN_B.tsv")
recount.b.df <- readr::read_tsv(recount.b.file)
```
```{r}
# SLE patients only
ifn.b.df <- recount.b.df %>%
  dplyr::filter(grepl("SLE patient", Patient))

ifn.b.df <- CalculateChangeFromBaseline(df = as.data.frame(ifn.b.df),
                                        variable.id = "LV",
                                        value.id = "Value")
df.file <- file.path(results.dir, "E-GEOD-39088_IFNk_recount2_PLIER_change.tsv")
readr::write_tsv(ifn.b.df, df.file)
```
```{r}
# plot
plot.file <- file.path(plot.dir, "E-GEOD-39088_IFNk_recount2_PLIER_change.pdf")
PlotChangeFromBaseline(df = ifn.b.df,
                       y.label = "Change in LV value",
                       plot.title = "recount2 PLIER - IFN-K treatment",
                       plot.path = plot.file,
                       facets = "LV ~ Day")
```
```{r}
rm(list = setdiff(ls(), c("%>%", "results.dir", "plot.dir", 
                          "data.prep.dir")))
```

## AMG 811 

**E-GEOD-78193; Welcher, et al. 2015.**

### Plotting Function

We'll want to generate a boxplot the interaction between disease state (e.g.,
healthy, SLE) and time point (day of trial) for each of the three methods.
We'll write a custom plotting function for this.

```{r}
PlotInteraction <- function(df, y.var, wrap.var, y.label, plot.title,
                            plot.subtitle = "Welcher, et al.") {
  # Given a data.frame that contains a "summary" expression level of some kind
  # for E-GEOD-78193 samples, make a boxplot where x/groups are 
  # interaction(Disease state, Day).  Not intended for use outside this 
  # context (& env)!
  #
  # Args:
  #   df: a (long form) data.frame containing the measurements for E-GEOD-78193
  #   y.var: variable used as y.var (string; evaluated with ggplot2::aes_string)
  #   wrap.var: string passed to ggplot2::facet_wrap; used for multiple LVs 
  #             or modules
  #   y.label: string, label for y-axis 
  #   plot.title: string, plot title
  #   plot.subtitle: string, plot subtitle; default "Welcher, et al."
  #
  # Returns:
  #   ggplot2::ggplot object
  
  ggplot2::ggplot(df, 
                  ggplot2::aes(x = interaction(`Disease state`, 
                                           `Day`), 
                               fill = interaction(`Disease state`, 
                                                  `Day`))) + 
    ggplot2::geom_boxplot(ggplot2::aes_string(y = y.var)) + 
    ggplot2::geom_point(ggplot2::aes_string(y = y.var),
                        alpha = 0.3, position = "jitter") +
    ggplot2::facet_wrap(as.formula(paste("~", wrap.var))) +
    ggplot2::theme_bw() + 
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                   legend.position = "none",
                   text = ggplot2::element_text(size = 15)) +
    ggplot2::labs(x = "interaction(Disease State, Day)",
                  y = y.label,
                  title = plot.title,
                  subtitle = plot.subtitle) +
    ggplot2::scale_fill_manual(values = c("seagreen3", "#deebf7", "#9ecae1",
                                          "#3182bd", "white")) +
    ggplot2::scale_x_discrete(labels = c("healthy", "SLE baseline", 
                                         "SLE day 15", "SLE day 56", 
                                         "SLE EOS"))
}

```

### Modular transcriptional analyses

```{r}
mod.file <- file.path(data.prep.dir,
                      "E-GEOD-78193_Chiche_et_al_module.tsv")
mod.summary.df <- readr::read_tsv(mod.file)
```

```{r}
p <- PlotInteraction(df = mod.summary.df,
                     y.var = "Summary",
                     wrap.var = "Module",
                     y.label = "Mean expression of genes in module (per sample)",
                     plot.title = "IFN Modular Framework Expression - Treatment with AMG 811")

plot.file <- file.path(plot.dir, "E-GEOD-78193_Chiche_et_al_boxplot.pdf")
ggplot2::ggsave(plot.file, plot = p, width = 11, height = 7, units = "in")
```

```{r}
rm(list = setdiff(ls(), c("%>%", "results.dir", "plot.dir", 
                          "data.prep.dir", "PlotInteraction")))
```

### PLIER trained on SLE WB compendium

```{r}
sle.b.file <- file.path(data.prep.dir, "E-GEOD-78193_SLE-WB_PLIER_IFN_B.tsv")
sle.b.df <- readr::read_tsv(sle.b.file)
```

```{r}
p <- PlotInteraction(df = sle.b.df,
                     y.var = "Value",
                     wrap.var = "LV",
                     y.label = "LV value",
                     plot.title = "SLE WB PLIER - Treatment with AMG 811")

plot.file <- file.path(plot.dir, "E-GEOD-78193_SLE_PLIER_boxplot.pdf")
ggplot2::ggsave(plot.file, plot = p, width = 11, height = 7, units = "in")
```

```{r}
rm(list = setdiff(ls(), c("%>%", "results.dir", "plot.dir", 
                          "data.prep.dir", "PlotInteraction")))
```

### PLIER trained on `recount2`

```{r}
recount.b.file <- file.path(data.prep.dir, 
                            "E-GEOD-78193_recount2_PLIER_IFN_B.tsv")
recount.b.df <- readr::read_tsv(recount.b.file)
```
```{r}
p <- PlotInteraction(df = recount.b.df,
                     y.var = "Value",
                     wrap.var = "LV",
                     y.label = "LV value",
                     plot.title = "recount2 PLIER - Treatment with AMG 811")

plot.file <- file.path(plot.dir, "E-GEOD-78193_recount2_PLIER_boxplot.pdf")
ggplot2::ggsave(plot.file, plot = p, width = 11, height = 7, units = "in")
```
