---
title: "Train a PLIER model on the NARES data"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

The NARES study was published in the following article:

> Grayson PC, Steiling K, Platt M, et al. [Defining the Nasal Transcriptome in Granulomatosis with Polyangiitis](https://dx.doi.org/10.1002/art.39185). _Arthritis & Rheumatology_, 2015. doi: 10.1002/art.39185.

We processed this data in the [`greenelab/rheum-data-plier`](https://github.com/greenelab/rheum-plier-data) 
repo.

This dataset (`n = 76`) contains nasal brushing samples from patients with 
granulomatosis with polyangiitis (GPA), a form of ANCA-associated vasculitis. 
Patients with GPA either had active, a prior history of, or no history of 
nasal disease.
A composite comparator group of patients with sarcoidosis, patients with
allergic rhinitis and healthy controls is also included.

Here we'll train a PLIER model and explore:

  * The sparsity of the U matrix
  * How many LVs are recovered? How many pathways?
  * What are the pathways that are represented?

## Libraries

```{r}
library(AnnotationDbi)
library(PLIER)
```

## NARES data

```{r}
nares.data <- readr::read_tsv(file.path("data", "expression_data", 
                                         "NARES_SCANfast_ComBat.pcl"))

symbol.obj <- org.Hs.eg.db::org.Hs.egSYMBOL
mapped.genes <- AnnotationDbi::mappedkeys(symbol.obj)
symbol.list <- as.list(symbol.obj[mapped.genes])
symbol.df <- as.data.frame(cbind(names(symbol.list), unlist(symbol.list)))
colnames(symbol.df) <- c("EntrezID", "GeneSymbol")

# get gene column name to match to facilitate use with dplyr
colnames(nares.data)[1] <- "EntrezID"

# matching types
symbol.df$EntrezID <- as.integer(as.character(symbol.df$EntrezID))

# inner join
annot.nares.data <- dplyr::inner_join(symbol.df, nares.data, by = "EntrezID")

symbol.file <- 
  file.path("data", "expression_data", 
            "NARES_SCANfast_ComBat_with_GeneSymbol.pcl")

readr::write_delim(annot.nares.data, path = symbol.file, delim = "\t")

# only leave the matrix with gene symbol as rownames
exprs.mat <- dplyr::select(annot.nares.data, -EntrezID)
rownames(exprs.mat) <- exprs.mat$GeneSymbol
exprs.mat <- as.matrix(dplyr::select(exprs.mat, -GeneSymbol))
rm(list = setdiff(ls(), c("exprs.mat")))
```

## Functions and directory set up

Organized this way for convenience w.r.t. removing objects from the environment
after gene identifier conversion.

```{r}
`%>%` <- dplyr::`%>%`
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "12")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "12")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## PLIER model

### Training

```{r}
# train model and save to file
nares.plier <- PLIERNewData(exprs.mat)
plier.file <- file.path(results.dir, "NARES_PLIER_model.RDS")
saveRDS(nares.plier, file = plier.file)

# save summary data.frame to results
summary.file <- file.path(results.dir, "NARES_summary.tsv")
readr::write_tsv(nares.plier$summary, path = summary.file)
```

### Number of LVs

```{r}
nrow(nares.plier$B)
```

The model learns 34 latent variables. 
From the trace during training, we know that 15 of them have `AUC>0.70`.

### Examine U and pathway coverage

As noted in the SLE WB PLIER notebook:

> `U` is the prior information coefficient matrix; it tells us how 
the prior information in the form of pathways/gene sets relates to LVs.

```{r}
PLIER::plotU(nares.plier, fontsize_row = 7)
```

To summarize *some* of these results:

* A latent variable associated with extracellular matrix formation is captured. 
`LV14` is associated with 
[`REACTOME_EXTRACELLULAR_MATRIX_ORGANIZATION`](http://software.broadinstitute.org/gsea/msigdb/cards/REACTOME_EXTRACELLULAR_MATRIX_ORGANIZATION.html), 
[`REACTOME_COLLAGEN_FORMATION`](http://software.broadinstitute.org/gsea/msigdb/cards/REACTOME_COLLAGEN_FORMATION.html),
and [`PID_INTEGRIN1_PATHWAY`](http://software.broadinstitute.org/gsea/msigdb/cards/PID_INTEGRIN1_PATHWAY.html).
* There's a neutrophil-associated LV, `LV3`.
* Interferon signaling is probably captured by `LV25`, but two immune cell 
types, M1 macrophages and CD8 T cells, are also associated with this LV.
This is likely due to their production of or activation by IFN and we saw a bit
of this overlap with cell types in other models (SLE WB, recount2).

```{r}
# save U plot to file
pdf(file.path(plot.dir, "NARES_U_plot.pdf"), height = 5, width = 7)
PLIER::plotU(nares.plier, fontsize_row = 8, fontsize_col = 8)
dev.off()
```

#### What proportion of the pathways input into the PLIER model are significantly associated with at least one LV?

```{r}
coverage.results <- GetPathwayCoverage(nares.plier)  # FDR < 0.05 by default
coverage.results$pathway
```

Less than 10% of the pathways we are interested in are significantly associated
with a latent variable.

#### How sparse is U?

Take into account _all_ pathways
```{r}
u.sparsity.all <- CalculateUSparsity(nares.plier,
                                     significant.only = FALSE)
ggplot2::ggplot(as.data.frame(u.sparsity.all),
                ggplot2::aes(x = u.sparsity.all)) +
  ggplot2::geom_density(fill = "blue", alpha = 0.5) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "proportion of positive entries in U",
                title = "All LVs") +
  ggplot2::theme(text = ggplot2::element_text(size = 15))
```

```{r}
summary(u.sparsity.all)
```

```{r}
plot.file <- file.path(plot.dir, "NARES_U_sparsity_all.png")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot())
```

What proportion of entries in the U matrix for each LV are significantly
associated with that LV?

```{r}
u.sparsity.sig <- CalculateUSparsity(nares.plier,
                                     significant.only = TRUE)
ggplot2::ggplot(as.data.frame(u.sparsity.sig),
                ggplot2::aes(x = u.sparsity.sig)) +
  ggplot2::geom_density(fill = "blue", alpha = 0.5) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "proportion of positive entries in U",
                title = "Pathway-associated LVs")  +
  ggplot2::theme(text = ggplot2::element_text(size = 15))
```

```{r}
summary(u.sparsity.sig)
```

```{r}
plot.file <- file.path(plot.dir, "NARES_U_sparsity_significant.png")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot())
```
