---
title: "Evaluating the recount2 subsampled models"
output: html_notebook
---

**J. Taroni 2018**

In `11-subsample_recount_PLIER.R`, we subsampled the recount2 dataset such that
it contained the same number of samples as the SLE WB compendium (`n = 1640`) 
ten times. 
We trained a PLIER model on each of the ten randomly selected datasets. 

Here, we'll evaluate the ten models in the following ways:

* Sparsity of `U` (prior information coefficient matrix; proxy for "ease 
of interpretation")
* Number of latent variables
* Pathway coverage (e.g., what percentage of pathways are associated with an 
LV, how many LVs have a pathway significantly associated with them)

## Functions and directory set up

```{r}
`%>%` <- dplyr::`%>%`
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "15")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "15")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Main evaluation

```{r}
# directory where the models RDS were saved
subsampled.dir <- file.path("results", "11")

# list files in the directory -- we'll us lapply to generate a list of list
plier.files <- list.files(subsampled.dir, full.names = TRUE)
```

```{r}
# read in models -- each of the files contains a list where PLIER corresponds
# to the PLIER model
model.list <- lapply(plier.files, function(x) readRDS(x)$PLIER)
names(model.list) <- sub(".RDS", "", sub(".*\\/", "", plier.files))
# evaluate models with wrapper function 
eval.list <- lapply(model.list, EvalWrapper)
```

```{r}
# reshape list to data.frame for wrangling
eval.df <- reshape2::melt(eval.list)
colnames(eval.df) <- c("value", "pathway_coverage_type", "metric", "model")

# U sparsity -- we'll keep all and significant only in the same data.frame
sparsity.df <- eval.df %>%
                dplyr::filter(metric %in% c("all.sparsity", "sig.sparsity")) %>%
                dplyr::mutate(sparsity_type = metric) %>%
                dplyr::select(c(model, sparsity_type, value))

# number of lvs
num.lvs.df <- eval.df %>%
                dplyr::filter(metric == "num.lvs") %>%
                dplyr::mutate(num_lvs = value) %>%
                dplyr::select(c(model, num_lvs))

# pathway coverage
pathway.df <- eval.df %>%
                dplyr::filter(metric == "pathway.coverage") %>% 
                dplyr::select(c(model, pathway_coverage_type, value))
```

```{r}
# write to file
sparsity.file <- file.path(results.dir, "subsampled_sparsity.tsv")
readr::write_tsv(sparsity.df, sparsity.file)

num.file <- file.path(results.dir, "subsampled_num_lvs.tsv")
readr::write_tsv(num.lvs.df, num.file)

pathway.file <- file.path(results.dir, "subsampled_pathway.tsv")
readr::write_tsv(pathway.df, pathway.file)
```

