---
title: "NARES differential expression"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

The NARES study ([Grayson, et al. _Arthritis Rheumatol_. 2015](https://dx.doi.org/10.1002/art.39185)) 
included three sets of patients with ANCA-associated vasculitis: those with 
active nasal disease, prior nasal disease, and those with no history of nasal 
disease.
There is also a composite comparator group.

Here, we'll test for PLIER LVs that are differentially expressed between groups.
We'll focus on the multiPLIER LVs (e.g., the recount2 PLIER model), as 
we have demonstrated the validity of this approach in prior analyses.

## Functions and directory set up

```{r}
`%>%` <- dplyr::`%>%` 
source(file.path("util", "test_LV_differences.R"))
```

```{r}
LVTestWrapper <- function(b.matrix,
                          sample.info.df,
                          phenotype.col,
                          file.lead,
                          plot.dir = "plots",
                          results.dir = "results",
                          use.bonferroni = FALSE,
                          significant.only = FALSE,
                          sig.threshold = 0.05) {
  # A wrapper function for TestLVDifferences and BoxplotDiffLV; does the 
  # reshaping required for plotting. Produces the following files: 1) tsv of the
  # differential expression results 2) a long form of the B matrix joined with
  # the sample information (sample.info.df) and 3) a PDF of boxplots
  # 
  # Args:
  #   b.matrix: a B matrix (latent variables are rows, samples are columns) 
  #             from a PLIER model. Can be the B element of the list 
  #             returned by PLIER::PLIER or the output of GetNewDataB
  #   sample.info.df: a long form data.frame that contains sample information,
  #                   sample names that match the B matrix sample identifiers
  #                   must be in a "Sample" column & it also must contain
  #                   the factor to group by for testing differential expression
  #                   (DE) and plotting
  #   phenotype.col: the column name of the column in sample.info.df to be used
  #                  for DE and plotting; character
  #   file.lead: string that designates the "beginning" of filenames
  #   plot.dir: plot directory where the boxplots PDF should be saved
  #   results.dir: results directory for DE results and reshaped B data.frame
  #   use.bonferroni: logical - should bonferroni correction be used for DE? 
  #                   if FALSE (default), will use "BH"
  #   significant.only: logical - should only differentially expressed LVs be
  #                     plotted? if FALSE (default), all will be plotted
  #   sig.threshold: the adj. P cutoff to be used if only plotting significant 
  #                  results; default is 0.05
  #                  
  # Returns:
  #   A list with the following elements
  #       limma: the results from TestLVDifferences 
  #       b.df: sample.b.df, prior to any filtering for plotting (if applicable)
  # 
  #   the following files are written by this function (see above):
  #     1) <results.dir>/<file.lead>_LV_limma_results.tsv
  #     2) <results.dir>/<file.lead>_B_long_sample_info.tsv
  #     3) <plot.dir>/<file.lead>_LV_boxplots.pdf
  
  `%>%` <- dplyr::`%>%`
  
  # error-handling
  # we need to join by "Sample" column for boxplots
  if (!("Sample" %in% colnames(sample.info.df))) {
    stop("'Sample' must be a column in sample.info.df")
  }
  # phenotype.col needs to be in column names
  if (!(phenotype.col %in% colnames(sample.info.df))) {
    stop("phenotype.col must be a column name in sample.info.df")
  }
  
  # initialize list to hold results to be returned
  return.list <- list()

  #### Differential Expression ####
  # get the named vector to use as the phenotype for testing differential 
  # expression
  phenotype.vector <- as.factor(make.names(sample.info.df[[phenotype.col]]))
  names(phenotype.vector) <- sample.info.df$Sample
  # test itself
  limma.df <- TestLVDifferences(b.matrix = b.matrix,
                                phenotype = phenotype.vector,
                                use.bonferroni = use.bonferroni)
  # write to file
  dlve.file <- file.path(results.dir, 
                         paste0(file.lead, "_LV_limma_results.tsv"))
  readr::write_tsv(limma.df, path = dlve.file)
  # add to list to be returned
  return.list[["limma"]] <- limma.df
  
  #### Reshape & join with sample information ####
  b.df <- reshape2::melt(b.matrix)
  colnames(b.df) <- c("LV", "Sample", "Value")
  sample.b.df <- dplyr::inner_join(b.df, sample.info.df, by = "Sample")
  long.file <- file.path(results.dir, 
                         paste0(file.lead, "_B_long_sample_info.tsv"))
  readr::write_tsv(sample.b.df, long.file)
  # add to list to be returned
  return.list[["b.df"]] <- sample.b.df
  
  #### Plotting ####
  plot.file <- file.path(plot.dir,
                         paste0(file.lead, "_LV_boxplots.pdf"))
  
  # if we only want significant LVs plotted, filter sample.b.df using the
  # adj.P.Val cutoff sig.threshold
  if (significant.only) {
    sig.lvs <- limma.df$LV[which(limma.df$adj.P.Val < sig.threshold)]
    sample.b.df <- sample.b.df %>%
      dplyr::filter(LV %in% sig.lvs)
  }
  
  BoxplotDiffLV(tidy.b.df = sample.b.df, 
                phenotype.column = phenotype.col,
                pdf.path = plot.file)
  

  return(return.list)
}
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "18")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "18")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

### Clinical data

This clinical data includes information like measures of disease severity 
([`BVAS`](https://doi.org/10.1136/ard.2008.101279)) and disease duration.
We're most interested in the disease labels (`Classification`).

```{r}
nares.demo <- readr::read_tsv(file.path("data", "sample_info", 
                                        "NARES_demographic_data.tsv")) %>%
  dplyr::mutate(Classification = dplyr::recode(Classification, 
                                               C1 = "Healthy",
                                               C2 = "Sarcoidosis",
                                               C3 = "Allergic Rhinitis",
                                               C4 = "EGPA",
                                               V1 = "GPA (no prior nasal disease)",
                                               V2 = "GPA with prior nasal disease",
                                               V3 = "GPA with active nasal disease"))
```

### recount2 LVs

```{r}
recount.b <- readRDS(file.path("results", "13", "NARES_recount2_B.RDS"))
```

## NARES differential latent variable expression (DLVE)

We're going to compare the seven classification groups, which correspond to
different disease labels (and were recoded above).

```{r}
nares.df.list <- LVTestWrapper(b.matrix = recount.b,
                                sample.info.df = nares.demo,
                                phenotype.col = "Classification",
                                file.lead = "NARES_recount2_model", 
                                plot.dir = plot.dir,
                                results.dir = results.dir)
```

### Example LV

```{r}
nares.limma.df <- nares.df.list$limma
nares.limma.df %>% 
  dplyr::filter(adj.P.Val < 0.05) %>% 
  dplyr::arrange(adj.P.Val)
```

```{r}
# read in recount2 summary data.frame
summary.file <- file.path("results", "08", "recount2_PLIER_summary.tsv")
recount.summary <- readr::read_tsv(summary.file)
```

Let's plot `LV96`, as it looks biologically interesting because of it's 
putative association with natural killer (NK) cells. 
We'd expect myeloid cell, specifically granulocyte, signatures to be 
differentially expressed in this dataset, but not necessarily lymphocyte 
signatures.
First, does the _name_ agree with the recount2 PLIER summary?

```{r}
recount.summary %>%
  dplyr::filter(`LV index` == 96)
```

Yes, `SVM NK cells activated` is significantly associated with `LV96`.

```{r}
recount.b.df <- nares.df.list$b.df
recount.b.df %>%
  dplyr::filter(LV == "96,SVM NK cells activated") %>%
  ggplot2::ggplot(ggplot2::aes(x = Classification, y = Value)) +
  ggplot2::geom_boxplot() +
  ggplot2::geom_jitter(position = ggplot2::position_jitter(0.15)) + 
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")) +
  ggplot2::ggtitle("LV 96 SVM NK cells activated")
```

