---
title: "GPA blood differential expression"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

Here, we'll perform differential expression (of LV) analyses for the PBMC data 
from Cheadle C, Berger AE, Andrade F, et al. 
[Transcription of PR3 and Related Myelopoiesis Genes in Peripheral Blood Mononuclear Cells in Active Wegener’s Granulomatosis](https://dx.doi.org/10.1002/art.27398). 
_Arthritis & Rheumatism_, 2010. doi: 10.1002/art.27398.

We'll be comparing the healthy controls, the patients with a GPA signature 
(GPA-positive) and the patients without a GPA signature (GPA-negative).
We'll first need to project this data into the recount2 PLIER model latent space.

## Functions and directory set up

```{r}
`%>%` <- dplyr::`%>%` 
source(file.path("util", "test_LV_differences.R"))
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "19")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "19")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

The GPA blood file is in the form of a GEO series matrix, which means the 
sample metadata is in what are essentially comment fields at the beginning of 
the file.
Thus, some ugliness/wrangling will be necessary.

### Expression data

```{r}
series.mat.file <- file.path("data", "expression_data", 
                             "GSE18885_series_matrix.txt")

# expression matrix
gpa.ma.data <- 
  readr::read_delim(series.mat.file, 
                    delim = "\t", 
                    comment = "!",
                    col_names = TRUE, 
                    skip = 1)

# this information about GPL6140 was downloaded on 15 Sept 2017, and reportedly
# last updated on 18 Jan 2013
gpl.from.geo <- readr::read_tsv(file.path("data", "expression_data", 
                                          "GPL6104-11576.txt"), comment = "#")

# use the gene symbols from the GPL file -- PLIER uses gene symbols
gpa.ma.annot <- gpl.from.geo %>%
  dplyr::select(c(ID, ILMN_Gene)) %>%
  dplyr::left_join(y = gpa.ma.data,
                   by = c("ID" = "ID_REF"))
colnames(gpa.ma.annot)[2] <- "Gene"
```

```{r}
# are there any duplicates in the gene symbol column?
sum(duplicated(gpa.ma.annot$Gene))
```

```{r}
# aggregate & write to file
agg.ma.df <- PrepExpressionDF(dplyr::select(gpa.ma.annot, -ID))
readr::write_tsv(agg.ma.df,
                 path = file.path("data", "expression_data",
                                  "GSE18885_annotated_mean.pcl"))

# get expression matrix
exprs.mat <- as.matrix(dplyr::select(agg.ma.df, -Gene))
rownames(exprs.mat) <- agg.ma.df$Gene
```

### Metadata

As mentioned above, metadata is also extracted from the series matrix file.

```{r}
# get more information from the series matrix file
# line 41 has the sample names
conn <- file(series.mat.file)
open(conn)
smpl.name <- read.table(conn, skip = 41, nrow = 1)
close(conn)

# line 51 contains the WG signature status -- I'll call this GPA signature
# GPA-positive, GPA-negative to be in line with the current disease name
conn <- file(series.mat.file)
open(conn)
gpa.sig.status <- read.table(conn, skip = 51, nrow = 1)
close(conn)

# this is the GEO sample accession information
conn <- file(series.mat.file)
open(conn)
gsm.info <- read.table(conn, skip = 77, nrow = 1)
close(conn)

# get those lines into data.frame format
smpl.info.df <- as.data.frame(t(dplyr::bind_rows(gsm.info, 
                                                 smpl.name, 
                                                 gpa.sig.status))[-1, ])
colnames(smpl.info.df) <- c("Sample", "Name", "GPA_signature")

# remove 1 row data.frames read in from the series matrix
rm(conn, gsm.info, smpl.name, gpa.sig.status)
```

```{r}
# extract fraction ("cell type(s)") from sample names
smpl.info.df <- smpl.info.df %>%
                  dplyr::mutate(Cell_type = sub(".*\\ ", "", Name))

# recode the GPA signature bit
smpl.info.df <-
  smpl.info.df %>%
  dplyr::mutate(GPA_signature = dplyr::case_when(
                  grepl("control", GPA_signature) ~ "Control", 
                  grepl("WG Sig -", GPA_signature) ~ "GPAneg",
                  grepl("WG Sig +", GPA_signature) ~ "GPApos"
                ))

# write to file
readr::write_tsv(smpl.info.df, 
                 path = file.path("data", "sample_info",
                                  "GSE18885_sample_info.tsv"))
```

```{r}
# remove data.frames that won't be needed
rm(agg.ma.df, gpa.ma.annot, gpa.ma.data, gpl.from.geo)
```

## Apply recount2 model

```{r}
# load model itself
recount.plier <- readRDS(file.path("data", "recount2_PLIER_data", 
                                   "recount_PLIER_model.RDS"))
# project
recount.b <- GetNewDataB(exprs.mat = as.matrix(exprs.mat),
                         plier.model = recount.plier)
```

```{r}
# save B matrix to file
b.file <- file.path(results.dir, "GPA_blood_recount2_B.RDS")
saveRDS(recount.b, file = b.file)
```

## Differential expression

### PBMC only

We're going to restrict our analyses to the peripheral blood mononuclear cell 
(PBMC) fraction.

```{r}
# sample information
pbmc.df <- smpl.info.df %>%
  dplyr::filter(Cell_type == "PBMC")
```

```{r}
# b matrix (latent variables)
pbmc.b <- recount.b[, pbmc.df$Sample]
```

### Differential expression analysis itself

```{r}
# get the phenotype labels -- we'll use GPA signature status
gpa.sig.vector <- pbmc.df$GPA_signature
names(gpa.sig.vector) <- pbmc.df$Sample
```

```{r}
# tests for differential expression
limma.df <- TestLVDifferences(b.matrix = pbmc.b,
                              phenotype = gpa.sig.vector)
dlve.file <- file.path(results.dir, 
                       "GPA_blood_recount2_model_LV_limma_results.tsv")
readr::write_tsv(limma.df, path = dlve.file)
```

```{r}
# plotting
# first, long format
pbmc.b.df <- reshape2::melt(pbmc.b)
colnames(pbmc.b.df) <- c("LV", "Sample", "Value")
# join with sample metadata
pbmc.b.df <- dplyr::inner_join(pbmc.df, pbmc.b.df, by = "Sample")
tidy.file <- file.path(results.dir, "GPA_blood_recount2_B_tidy_PBMC_only.tsv")
readr::write_tsv(pbmc.b.df, tidy.file)

# box plots
boxplot.file <- file.path(plot.dir, "GPA_blood_recount2_model_LV_boxplots.pdf")
BoxplotDiffLV(tidy.b.df = pbmc.b.df,
              phenotype.column = "GPA_signature",
              pdf.path = boxplot.file)
```
