---
title: "Top differentially expressed LVs in GPA blood"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

Explore the top differentially expressed LVs in the GPA blood dataset (PBMCs).
Are they pathways we could have captured with a PLIER model fit to this dataset?

## Functions and directory setup

```{r}
`%>%` <- dplyr::`%>%`
source(file.path("util", "plier_util.R"))
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "22")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "22")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Load DLVE results

```{r}
limma.file <- file.path("results", "19", 
                        "GPA_blood_recount2_model_LV_limma_results.tsv")
limma.df <- readr::read_tsv(limma.file)
```

Let's take a look at the top differentially expressed latent variables. 
We'll sort by FDR.

```{r}
limma.df %>%
  dplyr::filter(adj.P.Val < 0.05) %>%
  dplyr::arrange(adj.P.Val)
```

Let's explore `LV 599` a little further.

```{r}
recount.b.file <- file.path("results", "19", 
                            "GPA_blood_recount2_model_B_long_sample_info.tsv")
recount.b.df <- readr::read_tsv(recount.b.file) %>%
  dplyr::mutate(GPA_signature =
                  dplyr::case_when(
                    GPA_signature == "GPApos" ~ "GPA-positive",
                    GPA_signature == "GPAneg" ~ "GPA-negative",
                    TRUE ~ "Control"
                  ))
```

```{r}
recount.b.df %>%
  dplyr::filter(LV == "599,DMAP_ERY2") %>%
  ggplot2::ggplot(ggplot2::aes(x = GPA_signature, y = Value)) +
  ggplot2::geom_boxplot() +
  ggplot2::geom_jitter(position = ggplot2::position_jitter(0.2)) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "GPA signature", y = "LV 599",
                title = "Cheadle, et al. top LV") +
  ggplot2::theme(legend.position = "none",
                 text = ggplot2::element_text(size = 15),
                 plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"))
```

```{r}
plot.file <- file.path(plot.dir, "recount2_LV599_boxplot.pdf")
ggplot2::ggsave(plot.file, plot = ggplot2::last_plot())
```

## Load recount2 PLIER model

```{r}
recount.plier <- readRDS(file.path("data", "recount2_PLIER_data", 
                                   "recount_PLIER_model.RDS"))
```

### What are the pathways associated with `LV 599`?

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 599)
```

### What genes contribute to this LV (e.g., have the highest loadings)? 

Let's look at the top 50.

```{r}
head(sort(recount.plier$Z[, 599], decreasing = TRUE), 50)
```

Some of these are major or minor autoantigens in ANCA-associated vasculitis.
This is consistent with the original publication 
([Cheadle, et al. _A&R._ 2010.](https://doi.org/10.1002/art.27398)): 
the authors found that "neutrophil granule constituent genes" were highly 
expressed in the PBMC fraction from patients with GPA that could not be 
attributed to contamination with mature granulocytes.

#### Plot highlighting autoantigens

We'll use the list of autoantigens in Cheadle, et al.

```{r}
autoantigens <- c("MPO", "ELANE", "BPI", "CTSG", "LCN2", "AZU1", "PRTN3")
```

```{r}
# get into appropriate data.frame for bar plot
top.z.df <- as.data.frame(sort(recount.plier$Z[, 599],
                               decreasing = TRUE)[1:50])
top.z.df <- tibble::rownames_to_column(top.z.df, var = "Gene")
colnames(top.z.df)[2] <- "Z"

# add in autoantigen information
top.z.df <- top.z.df %>%
  dplyr::mutate(Autoantigen =
                  dplyr::case_when(
                    Gene %in% autoantigens ~ "Yes",
                    TRUE ~ "No"
                  ))

# reorder for plotting
top.z.df$Gene <- factor(top.z.df$Gene,
                        levels = top.z.df$Gene[50:1])
```

```{r}
p <- ggplot2::ggplot(top.z.df, 
                ggplot2::aes(x = Gene, y = Z, fill = Autoantigen)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::theme_bw() +
  ggplot2::scale_fill_manual(values = c("black", "red")) +
  ggplot2::coord_flip() +
  ggplot2::ggtitle("recount2 LV 599 Loadings") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, 
                                                    face = "bold"))
p + ggplot2::theme(axis.text = ggplot2::element_text(size = 6)) 
```

```{r}
plot.file <- file.path(plot.dir, "recount2_LV599_top_Z_bar.pdf")
ggplot2::ggsave(plot.file, 
                plot = p + 
                  ggplot2::theme(text = ggplot2::element_text(size = 15)), 
                width = 6, height = 8)
```

## GPA Blood-specific PLIER model

Would we have detected this "autoantigen signature" using a PLIER model
specifically trained on this GPA blood data?

```{r}
# remove objects we won't need for this next set of analyses
rm(limma.df, recount.b.df, limma.file, plot.file, recount.b.file,
   top.z.df)
```

### Read in expression data

```{r}
exprs.file <- file.path("data", "expression_data", 
                        "GSE18885_annotated_mean.pcl")
agg.ma.df <- readr::read_tsv(exprs.file)
# get expression matrix
exprs.mat <- as.matrix(dplyr::select(agg.ma.df, -Gene))
rownames(exprs.mat) <- agg.ma.df$Gene
```

### Train PLIER model

```{r}
gpa.plier <- PLIERNewData(exprs.mat)
```

```{r}
plier.file <- file.path(results.dir, "GPA_blood_PLIER_model.RDS")
saveRDS(gpa.plier, plier.file)
```

### Explore `U` matrix

Taking a look at the `U` (prior information coefficient) matrix will tell us
what kind of pathways were identified with this model.

```{r}
PLIER::plotU(gpa.plier, fontsize_row = 5, fontsize_col = 7)
```

```{r}
pdf(file.path(plot.dir, "GPA_blood_U_plot.pdf"))
PLIER::plotU(gpa.plier, fontsize_row = 6, fontsize_col = 7)
dev.off()
```

`LV13` and `LV20` look like the most likely candidates, but we'll look across
all 30 LVs.

### Does an "autoantigen LV" exist in the GPA blood PLIER model?

```{r}
GetTopNGenes <- function(z.vector, num.genes = 50){
  names(head(sort(z.vector, decreasing = TRUE), num.genes))
}

top.genes <- apply(gpa.plier$Z, 2, GetTopNGenes)
ag.logical <- apply(top.genes, 2, function(x) any(x %in% autoantigens))
which(ag.logical)
```

```{r}
head(sort(gpa.plier$Z[, 8], decreasing = TRUE), 50)
```

```{r}
# get into appropriate data.frame for bar plot
top.z.df <- as.data.frame(sort(gpa.plier$Z[, 8],
                               decreasing = TRUE)[1:50])
top.z.df <- tibble::rownames_to_column(top.z.df, var = "Gene")
colnames(top.z.df)[2] <- "Z"

# add in autoantigen information
top.z.df <- top.z.df %>%
  dplyr::mutate(Autoantigen =
                  dplyr::case_when(
                    Gene %in% autoantigens ~ "Yes",
                    TRUE ~ "No"
                  ))

# reorder for plotting
top.z.df$Gene <- factor(top.z.df$Gene,
                        levels = top.z.df$Gene[50:1])
```

```{r}
p <- ggplot2::ggplot(top.z.df, 
                ggplot2::aes(x = Gene, y = Z, fill = Autoantigen)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::theme_bw() +
  ggplot2::scale_fill_manual(values = c("black", "red")) +
  ggplot2::coord_flip() +
  ggplot2::ggtitle("GPA Blood LV 8 Loadings") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"))
p + ggplot2::theme(axis.text = ggplot2::element_text(size = 6)) 
```

```{r}
plot.file <- file.path(plot.dir, "GPA_blood_LV8_top_Z_bar.pdf")
ggplot2::ggsave(plot.file, 
                plot = p + 
                  ggplot2::theme(text = ggplot2::element_text(size = 15)), 
                width = 6, height = 8)
```

`GPA blood LV 8` may be _most_ similar to another recount2 LV like 
`recount2 LV 68`, which is also differentially expressed in this dataset.
Let's take a quick look at the top 50 genes for `recount2 LV 68`.

```{r}
head(sort(recount.plier$Z[, 68], decreasing = TRUE), 50)
```

_By eye_, this looks more similar to `GPA blood LV 8`, but let's do a more
formal analysis.

### Compare to recount2 LVs

```{r}
# get Z matrices from both models
recount.z <- as.data.frame(recount.plier$Z)
gpa.z <- as.data.frame(gpa.plier$Z)

# we'll need to add the gene identifiers (symbols in this case) to a column
# rather than as rownames -- this will facilitate joining
recount.z <- tibble::rownames_to_column(recount.z, var = "Gene")
colnames(recount.z)[2:ncol(recount.z)] <- paste0("recountLV", 
                                                 1:(ncol(recount.z) - 1))
gpa.z <- tibble::rownames_to_column(gpa.z, var = "Gene")
colnames(gpa.z)[2:ncol(gpa.z)] <- paste0("gpaLV", 1:(ncol(gpa.z) - 1))

# join -- only genes present in both models
z.df <- dplyr::inner_join(recount.z, gpa.z, by = "Gene")

# need matrix to calculate correlation
z.matrix <- as.matrix(z.df[, 2:ncol(z.df)])
rownames(z.matrix) <- z.df$Gene
```

```{r}
# calculate pearson correlation between LVs -- can we map between models using
# this distance metric?
cor.z.mat <- cor(z.matrix, method = "pearson")

# set diagonal to 0 to help find max correlation between LVs
diag(cor.z.mat) <- 0

# indices for each model
gpa.indx <- grep("gpa", rownames(cor.z.mat))
recount.indx <- grep("recount", rownames(cor.z.mat))

# pertinent indices
impt.cor.z.mat <- cor.z.mat[recount.indx, gpa.indx] 

# for each GPA blood model LV, want the highest correlated LV from recount
mapping.df <- reshape2::melt(impt.cor.z.mat, 
                              varnames = c("recount_LV", "GPA_blood_LV"),
                              value.name = "pearson_Z") %>%
                dplyr::group_by(GPA_blood_LV) %>%
                dplyr::top_n(1, pearson_Z)

mapping.df
```

It looks like `GPA blood LV 8` does in fact map to `recount2 LV 68` as 
determined by taking the Pearson correlation between loadings.

#### Heatmap of mapping

We'll select the top mapping LVs from the recount2 model as well as 
`recount2 LV 599` from `impt.cor.z.mat` for plotting as a heatmap.

```{r}
plot.index <- which(rownames(impt.cor.z.mat) %in% 
                      c(as.character(mapping.df$recount_LV), "recountLV599"))
heatmap.mat <- impt.cor.z.mat[plot.index, ]
pheatmap::pheatmap(heatmap.mat, main = "Z matrix mapping")
```

```{r}
plot.file <- file.path(plot.dir, "selected_cor_z_mat_heatmap.pdf")
pdf(plot.file)
pheatmap::pheatmap(heatmap.mat, main = "Z matrix mapping")
dev.off()
```

## Summary

Despite the lack of AAV expression data in the training compendium, the
recount2 PLIER model learns a latent variable highly relevant to AAV. 
Specifically, `recount2 LV 599` captures a granulocyte progenitor signature.
The top genes (as determined by the loadings, `Z`) are major (_PRTN3_, _MPO_) 
and minor (e.g., _AZU1_, _BPI_) autoantigens in ANCA-associated vasculitis.

We find an LV in the model specifically trained on the GPA blood dataset that
also captures the activity autoantigens (`GPA blood LV 8`).
However, the major autoantigens are not as highly ranked and this maps to 
another LV in the recount2 model, `recount2 LV 68`.
This suggests that the recount2 model (e.g., the multiPLIER approach) can 
identify additional, highly relevant LVs.
