---
title: "recount2 LVs in more severe or active ANCA-associated vasculitis"
output: html_notebook
---

**J. Taroni 2018**

In `21-AAV_DLVE`, we identified LVs that are differentially expressed in all
the ANCA-associated vasculitis datasets.
In this notebook, we'll do exploratory analyses of some of the recount2 PLIER
model LVs that look like they may be elevated in more severe or active AAV 
except for `LV793` looks to be consistently downregulated.

## Functions and directory set up

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "23")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "23")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

### recount2 PLIER model

```{r}
recount.plier <- readRDS(file.path("data", "recount2_PLIER_data",
                                   "recount_PLIER_model.RDS"))
```

## Explore LV10

Let's look at all the pathways associated with this LV.

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 10,
                FDR < 0.05)
```

```{r}
head(sort(recount.plier$Z[, 10], decreasing = TRUE), 50)
```

## Explore LV524

Let's look at all the pathways associated with this LV.

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 524,
                FDR < 0.05)
```

This is certainly capturing something about the myeloid lineage.
Recall that this LV was correlated with neutrophil count from the 
Banchereau, et al. dataset when we analyzed it in 
`07-sle_cell_type_recount2_model`.

Let's take a look at the genes contributing to this signal.

```{r}
head(sort(recount.plier$Z[, 524], decreasing = TRUE), 50)
```

## Explore LV624

No significant pathways.

```{r}
head(sort(recount.plier$Z[, 624], decreasing = TRUE), 50)
```

## Explore LV793

No significant pathways.

```{r}
head(sort(recount.plier$Z[, 793], decreasing = TRUE), 50)
```

## Explore LV937

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 937,
                FDR < 0.05)
```

```{r}
head(sort(recount.plier$Z[, 937], decreasing = TRUE), 50)
```