---
title: "recount2 LVs in more severe or active ANCA-associated vasculitis"
output: html_notebook
---

**J. Taroni 2018**

In `21-AAV_DLVE`, we identified LVs that are differentially expressed in all
the ANCA-associated vasculitis datasets.
In this notebook, we'll do exploratory analyses of some of the recount2 PLIER
model LVs that look like they show consistent directionality across the 3
tissues.

## Functions and directory set up

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`
```

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "23")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "23")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

### recount2 PLIER model

```{r}
recount.plier <- readRDS(file.path("data", "recount2_PLIER_data",
                                   "recount_PLIER_model.RDS"))
```

## Upregulated

LVs that are consistently upregulated in severe or active AAV as compared to
controls.

### Explore LV10

Let's look at all the pathways associated with this LV.

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 10,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 10], decreasing = TRUE))
```

### Explore LV53

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 53,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 53], decreasing = TRUE))
```

### Explore LV63

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 63,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 63], decreasing = TRUE))
```

### Explore LV524

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 524,
                FDR < 0.05)
```

This is certainly capturing something about the myeloid lineage.
Recall that this LV was correlated with neutrophil count from the 
Banchereau, et al. dataset when we analyzed it in 
`07-sle_cell_type_recount2_model`.

Let's take a look at the genes contributing to this signal.

```{r}
data.frame(loadings = sort(recount.plier$Z[, 524], decreasing = TRUE))
```

### Explore LV624

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 624,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 624], decreasing = TRUE))
```

### Explore LV937

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 937,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 937], decreasing = TRUE))
```

## Downregulated

LVs that are consistently downregulated in severe or active AAV as compared to
controls.

### Explore LV139

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 139,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 139], decreasing = TRUE))
```

### Explore LV332

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 332,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 332], decreasing = TRUE))
```

### Explore LV346

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 346,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 346], decreasing = TRUE))
```

### Explore LV631

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 631,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 631], decreasing = TRUE))
```

### Explore LV793

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 793,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 793], decreasing = TRUE))
```

### Explore LV887

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 887,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 887], decreasing = TRUE))
```

### Explore LV909

```{r}
recount.plier$summary %>%
  dplyr::filter(`LV index` == 909,
                FDR < 0.05)
```

```{r}
data.frame(loadings = sort(recount.plier$Z[, 909], decreasing = TRUE))
```

