---
title: "Describe recount2 compendium"
output: html_notebook
---

**J. Taroni 2018**

In response to comments, we wish to provide a bit more background information 
about the recount2 compendium.
To do so, we'll gather information from the following sources:

* The expression data used to train the model
* [MetaSRA](http://metasra.biostat.wisc.edu/index.html)
* SRAdb (for accession identifier conversion)

## Download MetaSRA data

```{sh}
wget -N -O data/sample_info/metasra.v1-4.json http://metasra.biostat.wisc.edu/static/metasra_versions/v1.4/metasra.v1-4.json
```

## Libraries, functions

```{r}
# BiocInstaller::biocLite("SRAdb")
# library(SRAdb)

# magrittr pipe
`%>%` <- dplyr::`%>%`
```

Custom function for converting between the sample (e.g., `SRS`) accession codes
and the sample names in recount2 which follow this pattern: `SRPxxxxxx.SRRxxxxx`

```{r}
ConvertToRecountSampleName <- function(sample.vector, conversion.df) {
  # Takes a vector of sample accession codes and converts them to sample/column
  # names for use with recount2 expression matrices
  # 
  # Args:
  #   sample.vector: a vector of sample accession codes
  #   conversion.df: a data.frame that provides the mappings between study,
  #                  sample, and run accession codes -- must have "run", 
  #                  "study", and "sample" column names
  # 
  # Returns:
  #   sample.names: a vector of sample/column names for use with recount2 --
  #                 these are the study and run accession codes separated with
  #                 a `.`
  
  # error handling
  check.colnames <- all(c("run", "study", "sample") %in% 
                            colnames(conversion.df))
  if (!check.colnames) {
    stop("colnames(conversion.df) must contain 'run', 'study', 'sample'")
  }
  
  # magrittr pipe, just in case
  `%>%` <- dplyr::`%>%`
  
  # filter to only the samples in the supplied vector and then concatenate their
  # study and sample accession codes
  sample.names.df <- conversion.df %>%
    dplyr::filter(`sample` %in% sample.vector) %>%
    dplyr::mutate(recount.name = paste(study, run, sep = "."))
 
  # return a vector of the recount2 names
  return(sample.names.df$recount.name) 
}
```

## Read in data

### recount2 data used to train PLIER

```{r}
plier.file <- file.path("data", "recount2_PLIER_data", 
                        "recount_data_prep_PLIER.RDS")
plier.data <- readRDS(plier.file)
```

The experiment accession is in the column name of the expression data matrix,
as is the run accession.

```{r}
experiment.accession <- unique(stringr::word(colnames(plier.data$rpkm.cm), 1, 
                                             sep = "[.]"))
head(experiment.accession)
```

```{r}
run.accession <- stringr::word(colnames(plier.data$rpkm.cm), 2, sep = "[.]")
```

```{r}
# remove the training data from the workspace, we no longer need it
rm(plier.data)
```

### MetaSRA

```{r}
metasra.file <- file.path("data", "sample_info", "metasra.v1-4.json")
metasra <- jsonlite::read_json(metasra.file)
```

The MetaSRA project uses _sample_ identifiers (`SRS`), whereas recount2 uses 
_study_ (`SRP`) and _run_ (`SRR`) identifiers.
We'll need to convert between the two. 
We'll use the `SRAdb` Bioconductor package to do so.

#### Identifier conversion

For notebook purposes, I will comment most of this out, as the file downloaded 
and unzipped by calling `SRAdb::getSRAdbFile()` is ~35GB.
I will save (& commit) the most important part: the data.frame that contains
the mapping between accessions.

```{r}
# sql.file <- getSRAdbFile()
```

```{r}
# sql.file <- "SRAmetadb.sqlite"
# sra.con <- dbConnect(SQLite(), sql.file)  
# given the run accession, return the study and sample accesions
# conversion.df <- sraConvert(run.accession, 
#                             out_type = c("study", "sample"), sra.con)
# conversion.file <- file.path("data/sample_info/recount2_srr_srs_srp.tsv")
# readr::write_tsv(conversion.df, conversion.file)
```

```{sh}
# md5sum SRAmetadb.sqlite
```

Because we've commented the above out for size concerns, we'll go ahead and 
read `conversion.df` back in using the TSV file we've saved and committed.

```{r}
conversion.file <- file.path("data", "sample_info", "recount2_srr_srs_srp.tsv")
conversion.df <- readr::read_tsv(conversion.file)
```

## Describe recount2

We'll also use these metadata to split the recount2 data into different training
sets based on the biological conditions or sample types.

```{r}
# only proceed with samples that are in the recount2 compendium that we used
filtered.metasra <- metasra[which(names(metasra) %in% conversion.df$sample)]
```

Not all of the runs in the subset of the recount2 compendium we used are 
present in MetaSRA.

```{r}
length(filtered.metasra) / length(run.accession)
```

But the majority are.

### Sample type

The MetaSRA project ([Bernstein, Doan, and Dewey. _Bioinformatics._ 2017.](https://doi.org/10.1093/bioinformatics/btx334)) 
predicted the "sample type" -- e.g., cell line vs. tissue.
What's the breakdown in the compendium we used for training (with no regard 
for _confidence_)?

```{r}
sample.type <- lapply(filtered.metasra, function(x) x$`sample type`)
table(unlist(sample.type))
```

The bulk of samples are predicted to be cell lines and tissue samples. 
Let's get a list of recount2 sample/column names for the cell line samples and
the tissue samples.

```{r}
cell.line.samples <- names(sample.type[which(sample.type == "cell line")])
cell.line.accessions <- ConvertToRecountSampleName(cell.line.samples,
                                                   conversion.df)
cell.line.file <- file.path("data", "sample_info", 
                            "recount2_cell_line_accessions.tsv")
write.table(cell.line.accessions, file = cell.line.file, sep = "\t", 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

```{r}
tissue.samples <- names(sample.type[which(sample.type == "cell line")])
tissue.accessions <- ConvertToRecountSampleName(tissue.samples,
                                                   conversion.df)
tissue.file <- file.path("data", "sample_info", 
                            "recount2_tissue_accessions.tsv")
write.table(tissue.accessions, file = tissue.file, sep = "\t", 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

### Experimental Factor Ontology (EFO)

We'll use the mapped ontology terms -- specifically, the mapped
[Experimental Factor Ontology](https://www.ebi.ac.uk/efo/) terms -- provided in
MetaSRA to get an idea of what's in our training set.

We've been asked specifically about autoimmune samples and cancer samples, so
let's start there.

### Autoimmune disease

```{r}
# "autoimmune disease"
# https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0005140
autoimmune.samples <- 
  lapply(filtered.metasra, 
         function(x) any("EFO:0005140" %in% x$`mapped ontology terms`))
sum(unlist(autoimmune.samples))
```

There are 99 SLE samples (and 18 controls) in the single SLE whole blood data 
set in recount2 
([`SRP062966`](http://trace.ncbi.nlm.nih.gov/Traces/sra/?study=SRP062966)).

### Cancer

```{r}
# "cancer"
# https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0000311
cancer.samples <- 
  lapply(filtered.metasra, 
         function(x) any("EFO:0000311" %in% x$`mapped ontology terms`))
sum(unlist(cancer.samples))
```

Let's write these accession codes to file.

```{r}
# just use the accession code, which is the name of the list for which the
# element == TRUE
cancer.samples <- names(cancer.samples[which(unlist(cancer.samples))])
cancer.accessions <- 
  ConvertToRecountSampleName(cancer.samples, 
                             conversion.df)
cancer.file <- file.path("data", "sample_info", 
                            "recount2_cancer_accessions.tsv")
write.table(cancer.accessions, file = cancer.file, sep = "\t", 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

### What about treated samples?

[`treatment`](https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0000727)

From EFO term:

> A process in which the act is intended to modify or alter some other 
material entity

```{r}
# "treatment"
# https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0000727
treatment.samples <- 
  lapply(filtered.metasra, 
         function(x) any("EFO:0000727" %in% x$`mapped ontology terms`))
sum(unlist(treatment.samples))
```

### Blood samples

We'll want to train models _exclusively_ on blood samples and on 
_everything-but-blood_ samples. 
We're interested in this because the model learns leukocyte associated
latent variables.

We'll use the [Uberon](https://www.ebi.ac.uk/ols/ontologies/uberon) term for this 
[`UBERON:0000178`](https://www.ebi.ac.uk/ols/ontologies/uberon/terms?short_form=UBERON_0000178)

```{r}
blood.samples <- 
  lapply(filtered.metasra, 
         function(x) any("UBERON:0000178" %in% x$`mapped ontology terms`))
sum(unlist(blood.samples))
```

Convert the blood samples and the everything-but-blood samples to the recount2
sample/column names and write to file.

```{r}
# just the accession codes
blood.samples <- names(blood.samples[which(unlist(blood.samples))])
blood.accessions <- ConvertToRecountSampleName(blood.samples, conversion.df)
blood.file <- file.path("data", "sample_info", 
                            "recount2_blood_accessions.tsv")
write.table(blood.accessions, file = blood.file, sep = "\t", 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

```{r}
not.blood.samples <- setdiff(names(filtered.metasra), blood.samples)
not.blood.accessions <- ConvertToRecountSampleName(not.blood.samples, 
                                                   conversion.df)
not.blood.file <- file.path("data", "sample_info", 
                            "recount2_other_tissues_accessions.tsv")
write.table(not.blood.accessions, file = not.blood.file, sep = "\t", 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

