---
title: "Describe recount2 compendium"
output: html_notebook
---

**J. Taroni 2018**

In response to comments, we wish to provide a bit more background information 
about the recount2 compendium.
To do so, we'll gather information from the following sources:

* The expression data used to train the model
* [MetaSRA](http://metasra.biostat.wisc.edu/index.html)
* SRAdb (for accession identifier conversion)

## Download MetaSRA data

```{sh}
wget -O data/sample_info/metasra.v1-4.json http://metasra.biostat.wisc.edu/static/metasra_versions/v1.4/metasra.v1-4.json
```

## Libraries, functions

```{r}
# BiocInstaller::biocLite("SRAdb")
library(SRAdb)

# magrittr pipe
`%>%` <- dplyr::`%>%`
```

## Read in data

### recount2 data used to train PLIER

```{r}
plier.file <- file.path("data", "recount2_PLIER_data", 
                        "recount_data_prep_PLIER.RDS")
plier.data <- readRDS(plier.file)
```

The experiment accession is in the column name of the expression data matrix,
as is the run accession.

```{r}
experiment.accession <- unique(stringr::word(colnames(plier.data$rpkm.cm), 1, 
                                             sep = "[.]"))
head(experiment.accession)
```

```{r}
run.accession <- stringr::word(colnames(plier.data$rpkm.cm), 2, sep = "[.]")
```

```{r}
# remove the training data from the workspace, we no longer need it
rm(plier.data)
```

### MetaSRA

```{r}
metasra.file <- file.path("data", "sample_info", "metasra.v1-4.json")
metasra <- jsonlite::read_json(metasra.file)
```

The MetaSRA project uses _sample_ identifiers (`SRS`), whereas recount2 uses 
_study_ (`SRP`) and _run_ (`SRR`) identifiers.
We'll need to convert between the two. 
We'll use the `SRAdb` Bioconductor package to do so.

#### Identifier conversion

For notebook purposes, I will comment most of this out, as the file downloaded 
and unzipped by calling `SRAdb::getSRAdbFile()` is ~35GB.
I will save (& commit) the most important part: the data.frame that contains
the mapping between accessions.

```{r}
# sql.file <- getSRAdbFile()
```

```{r}
# sql.file <- "SRAmetadb.sqlite"
# sra.con <- dbConnect(SQLite(), sql.file)  
# given the run accession, return the study and sample accesions
# conversion.df <- sraConvert(run.accession, 
#                             out_type = c("study", "sample"), sra.con)
# conversion.file <- file.path("data/sample_info/recount2_srr_srs_srp.tsv")
# readr::write_tsv(conversion.df, conversion.file)
```

```{sh}
# md5sum SRAmetadb.sqlite
```

## Describe recount2

```{r}
# only proceed with samples that are in the recount2 compendium that we used
filtered.metasra <- metasra[which(names(metasra) %in% conversion.df$sample)]
```

Not all of the runs in the subset of the recount2 compendium we used are 
present in MetaSRA.

```{r}
length(filtered.metasra) / length(run.accession)
```

But the majority are.

### Sample type

The MetaSRA project ([Bernstein, Doan, and Dewey. _Bioinformatics._ 2017.](https://doi.org/10.1093/bioinformatics/btx334)) 
predicted the "sample type" -- e.g., cell line vs. tissue.
What's the breakdown in the compendium we used for training (with no regard 
for _confidence_)?

```{r}
sample.type <- lapply(filtered.metasra, function(x) x$`sample type`)
table(unlist(sample.type))
```

The bulk of samples are predicted to be cell lines and tissue samples.

### Experimental Factor Ontology (EFO)

We'll use the mapped ontology terms -- specifically, the mapped
[Experimental Factor Ontology](https://www.ebi.ac.uk/efo/) terms -- provided in
MetaSRA to get an idea of what's in our training set.

We've been asked specifically about autoimmune samples and cancer samples, so
let's start there.

### Autoimmune disease

```{r}
# "autoimmune disease"
# https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0005140
autoimmune.samples <- 
  lapply(filtered.metasra, 
         function(x) any("EFO:0005140" %in% x$`mapped ontology terms`))
sum(unlist(autoimmune.samples))
```

There are 99 SLE samples (and 18 controls) in the single SLE whole blood data 
set in recount2 
([`SRP062966`](http://trace.ncbi.nlm.nih.gov/Traces/sra/?study=SRP062966)).

### Cancer

```{r}
# "cancer"
# https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0000311
cancer.samples <- 
  lapply(filtered.metasra, 
         function(x) any("EFO:0000311" %in% x$`mapped ontology terms`))
sum(unlist(cancer.samples))
```

### What about treated samples?

[`treatment`](https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0000727)

From EFO term:

> A process in which the act is intended to modify or alter some other 
material entity

```{r}
# "treatment"
# https://www.ebi.ac.uk/ols/ontologies/efo/terms?short_form=EFO_0000727
treatment.samples <- 
  lapply(filtered.metasra, 
         function(x) any("EFO:0000727" %in% x$`mapped ontology terms`))
sum(unlist(treatment.samples))
```

