---
title: "Evaluating subsampling: sample size 'sweep' and biological contexts"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

In `29-train_models_different_sample_size.sh`, we trained models of different 
sample sizes (5x repeats with different random seeds) and in 
`28-train_different_biological_contexts.sh`, we trained models using training
sets comprised of different (predicted by MetaSRA) 
Here, we're interested in evaluating these models.
Specifically, we want to know:

* **How many latent variables were learned by each model?** 
  We expect this number to increase with sample size.
* **Of the pathway supplied to the model during training, what proportion of 
  these are captured by the model?**
  (We refer to this measure as "pathway coverage.")
  We also expect this number to go up with sample size, at least initially, and
  for this metric to be somewhat stable across repeats, based on the results
  displayed/plotted in `17-plotting_repeat_evals` (calculated in 
  `15-evaluate_subsampling`).
  The different biological contexts datasets are of varying sizes as well.
* **What pathways are captured?**
  Here, again we're concerned with pathways that were supplied to the model 
  during training.
  We'll want to do this evaluation in a separate notebook, but we expect a model
  trained on something like blood will be quite good at capturing immune cell
  signatures.
* **What oncogenic pathways are captured by the model?**
  The [oncogenic pathways from MSigDB](software.broadinstitute.org/gsea/msigdb/genesets.jsp?collection=C6) 
  were _not_ supplied to the model during training, so this is a holdout set.
  I hypothesize that models trained on more samples and models that with 
  "more relevant" training sets (e.g., cancer) will do a bit better at
  capturing these heldout pathways.

## Set up

#### Custom functions + libraries

```{r}
`%>%` <- dplyr::`%>%`
source(file.path("util", "plier_util.R"))
```

```{r}
# intended for use only in this context -- functionalized because the two
# sets of models have the same structure (e.g., repeats are performed for 
# both the sample size evaluation _and_ the biological context evaluation)
EvalRepeats <- function(model.files, holdout.mat, outer.identifier, 
                        model.names) {
  # For a vector of model files that are deemed to be related in some way
  # (e.g., different sample sizes) AND that contain repeats (output of 
  # scripts/subsampling_PLIER.R with --num_repeats > 1), perform the 
  # evaluations contained in EvalWrapperWithHoldout AND tidy the results
  # for use downstream. Returns a list of data.frames.
  # 
  # Args:
  #  model.files: a vector of full paths to different, related model files
  #  holdout.mat: matrix of heldout pathways for use with EvalWrapperWithHoldout
  #               see the documentation of CalculateHoldoutAUC for more info
  #  outer.identifier: character; what are the different, related models 
  #                    evaulating? e.g., "sample_size" or "biological_context"
  #  model.names: what should the names of the list of results be? a character
  #               vector
  #              
  # Returns:
  #   a list of data.frames that contains the following elements: holdout.df,
  #   num.lvs.df, and coverage.df
  
  # calculate pathway coverage, number of latent variables, and the AUC
  # for the heldout pathways
  results.list <- 
    lapply(model.files, # for each set of models
           function(x) {
             plier.models <- readRDS(x)  # read in list of models
             # for each individual repeat
             lapply(plier.models, function(y) {
               EvalWrapperWithHoldout(plier.model = y$PLIER,
                                      holdout.matrix = holdout.mat)
             })
           })
  
  # we'll name the elements of the list based on the model.names argument
  names(results.list) <- model.names
  
  ## HELDOUT PATHWAYS ----------------------------------------------------------
  
  # extract the heldout results element -- the structure of the results list is
  # sample size -> repeat -> pathway.coverage, num.lvs, heldout.results
  # we want all the heldout.results from results.list
  holdout.results <- lapply(results.list, 
                            function(x) lapply(x, 
                                               function(y) y$heldout.results))
  
  # now we want the results in the form of a data.frame, where we include 
  # information about the sample size or biological context as well as the 
  # random seed used to generate that repeat's results
  holdout.df <- 
    dplyr::bind_rows(
      # for each condition, create a data.frame of all the holdout results and
      # include the random seed as an identifier
      lapply(holdout.results, 
             function(x) dplyr::bind_rows(x, .id = "seed")), 
      # use the outer.identifer arg as an id when we bind all the data.frames 
      # together
      .id = outer.identifier 
    )
  
  ## NUMBER OF LATENT VARIABLES ------------------------------------------------
  
  # first extract the num.lvs elements, then melt into a data.frame
  num.lvs.df <- 
    reshape2::melt(lapply(results.list,
                          function(x) lapply(x, 
                                             function(y) y$num.lvs)),
                   value.name = "number_of_latent_variables")
  colnames(num.lvs.df)[2:3] <- c("seed", outer.identifier)
  
  ### PATHWAY COVERAGE ---------------------------------------------------------
  
  # extract only the pathway.coverage elements
  pathway.coverage.list <- 
    lapply(results.list, 
           function(x) lapply(x, function(y) y$pathway.coverage))
  # melt into data.frame
  coverage.df <- reshape2::melt(pathway.coverage.list)
  # since we've melted from a list, rename the columns 
  colnames(coverage.df)[2:4] <- c("metric", "seed", outer.identifier)
  
  # we're only really interested in 2 out of 3 of the metrics calculated by
  # GetPathwayCoverage -- filter to only those and recode 
  coverage.df <- coverage.df %>%
    dplyr::filter(metric %in% c("pathway", "lv")) %>%
    dplyr::mutate(metric = 
                    dplyr::case_when(
                      (metric == "lv") ~ 
                        "LV associated with pathways",
                      (metric =="pathway") ~ "pathway coverage"
                    ))
 
  ## RETURN --------------------------------------------------------------------
  
  # return a list of wrangled data.frames
  return(list(
    holdout.df = holdout.df,
    num.lvs.df = num.lvs.df,
    coverage.df = coverage.df
  ))
   
}
```

```{r}
# this is required to get the oncogenic pathways -- we're going to use this
# as our holdout set here
library(PLIER)
```

#### Directory setup

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "30")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "30")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

#### Models to read in

All models we'll evaluate are in `models/`

```{r}
models.dir <- "models"
# we want all the models with 'subsampled' in the file name -- these are the
# sample size evaluations
size.model.files <- list.files(models.dir, pattern = "subsampled", 
                               full.names = TRUE)
size.model.files
```

Now for the different biological contexts

```{r}
# biological context models have 'accessions' in the file names
context.model.files <- list.files(models.dir, pattern = "accessions",
                                  full.names = TRUE)
context.model.files
```


## Evaluations

We're going to use the oncogenic pathways that come with PLIER as our holdout
set.

```{r}
# load in holdout set
data("oncogenicPathways")
```

### Sample size

```{r}
# we'll name the elements of the list based on their sample size, which we 
# can extract from the filenames themselves
size.model.names <- sub(".RDS", "", sub(".*[_]", "", size.model.files))
```

Main evaluation with `EvalRepeats`

```{r}
size.results <- EvalRepeats(model.files = size.model.files,
                            holdout.mat = oncogenicPathways,
                            outer.identifier = "sample_size",
                            model.names = size.model.names)
```

#### Heldout pathways

```{r}
# write to file
size.holdout.file <- file.path(results.dir, 
                               "subsampled_oncogenic_pathways_AUC.tsv")
readr::write_tsv(size.results$holdout.df, path = size.holdout.file)
```

#### Number of latent variables

```{r}
# write to file
size.num.lvs.file <- file.path(results.dir, "subsampled_number_of_lvs.tsv")
readr::write_tsv(size.results$num.lvs.df, path = size.num.lvs.file)
```

#### Pathway coverage

```{r}
# write to file!
size.coverage.file <- file.path(results.dir, "subsampled_pathway_coverage.tsv")
readr::write_tsv(size.results$coverage.df, path = size.coverage.file)
```

Remove `size.results` for memory reasons -- we've saved everything to file.

```{r}
rm(size.results)
```

### Biological contexts

```{r}
# extract the biological contexts from the file names 
context.model.names <- stringr::str_match(context.model.files, 
                                          "recount2_(.*?)_accessions")[, 2]
```

Main evaluation with `EvalRepeats`

```{r}
context.results <- EvalRepeats(model.files = context.model.files,
                               holdout.mat = oncogenicPathways,
                               outer.identifier = "biological_context",
                               model.names = context.model.names)
```

#### Heldout pathways

```{r}
# write to file
context.holdout.file <- 
  file.path(results.dir, "biological_contexts_oncogenic_pathways_AUC.tsv")
readr::write_tsv(context.results$holdout.df, path = context.holdout.file)
```

#### Number of latent variables

```{r}
# write to file
context.num.lvs.file <- 
  file.path(results.dir, "biological_context_number_of_lvs.tsv")
readr::write_tsv(context.results$num.lvs.df, path = context.num.lvs.file)
```

#### Pathway coverage

```{r}
# write to file!
context.coverage.file <- 
  file.path(results.dir, "biological_context_pathway_coverage.tsv")
readr::write_tsv(context.results$coverage.df, path = context.coverage.file)
```
