---
title: "DIPG: data cleaning"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

In this notebook, we'll clean diffuse intrinsic pontine glioma (DIPG) data.
There is no DIPG data in recount2, so this is another use case for this
project.

We'll be working with two datasets stored in the
[`greenelab/rheum-plier-data`](https://github.com/greenelab/rheum-plier-data) 
repository:

* [`E-GEOD-26576`](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-26576/)
* [`GSE50021`](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE50021)

**Citations:**

> Paugh BS, Broniscer A, Qu C, et al. [Genome-wide analyses identify recurrent 
amplifications of receptor tyrosine kinases and cell-cycle regulatory genes in 
diffuse intrinsic pontine glioma.](https://dx.doi.org/10.1200/JCO.2011.35.5677) 
_J Clin Oncol._ 2011;29(30):3999-4006.

> Buczkowicz P, Hoeman C, Rakopoulos P, et al. [Genomic analysis of diffuse 
intrinsic pontine gliomas identifies three molecular subgroups and recurrent 
activating _ACVR1_ mutations.](https://dx.doi.org/10.1038/ng.2936) _Nat Genet._ 
2014;46(5):451-6. 

## Set up

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`
# we need the function that aggregates duplicate gene identifiers to the
# mean value
source(file.path("util", "test_LV_differences.R"))
```

#### Directory setup

```{r}
# directory that holds the gene expression files
exprs.dir <- file.path("data", "expression_data")
# directory that holds the sample metadata
sample.info.dir <- file.path("data", "sample_info")
```

## Read in and clean data

### GSE50021

We have the series matrix for `GSE50021`, which contains both the expression
values and the metadata

```{r}
series.mat.file <- file.path(exprs.dir, "GSE50021_series_matrix.txt")

# expression matrix -- everything but the comment lines that begin with !
ma.data <- 
  readr::read_delim(series.mat.file, 
                    delim = "\t", 
                    comment = "!",
                    col_names = TRUE, 
                    skip = 1)
```

#### Gene identifier conversion

```{r}
# The GPL information from GEO, which was made public on Jul 18, 2011 and 
# last updated on Jan 18, 2013
gpl.info.df <- readr::read_tsv(file.path(exprs.dir, "GPL13938-11302.txt"),
                               comment = "#")
```

```{r}
annot.gse50021.df <- gpl.info.df %>%
  # from the GEO information, grab just the probe identifier and the gene
  # symbol columns
  dplyr::select(c(ID, Symbol)) %>%
  # only ILMN IDs (probes) in both
  dplyr::inner_join(ma.data, by = c("ID"  = "ID_REF")) %>%
  # collapsing duplicate symbols later will require the symbols to be in the
  # first column, called "Gene", with no additional columns
  dplyr::mutate(Gene = Symbol) %>%
  dplyr::select(-ID, -Symbol) %>%
  dplyr::select(Gene, dplyr::everything())
```

Collapse duplicate symbols and write to file.

```{r}
# summarize to mean
annot.mean.df <- PrepExpressionDF(annot.gse50021.df)
readr::write_tsv(annot.mean.df, file.path(exprs.dir, "GSE50021_mean_agg.pcl"))
```

#### Sample metadata

As mentioned above, metadata is also extracted from the series matrix file.
We do this for a single line at a time that we've picked based on the relevance
to any downstream analysis we might do (contact information, for example, 
does not help us in this context).

```{r}
# get more information from the series matrix file
# sample accession e.g., GSMXXXXX
conn <- file(series.mat.file)
open(conn)
sample.accession <- read.table(conn, skip = 79, nrow = 1)
close(conn)

# source name
conn <- file(series.mat.file)
open(conn)
source.name <- read.table(conn, skip = 85, nrow = 1)
close(conn)

# tissue
conn <- file(series.mat.file)
open(conn)
tissue <- read.table(conn, skip = 88, nrow = 1)
close(conn)

# gender
conn <- file(series.mat.file)
open(conn)
gender <- read.table(conn, skip = 89, nrow = 1)
close(conn)

# age at diagnosis
conn <- file(series.mat.file)
open(conn)
age.dx <- read.table(conn, skip = 90, nrow = 1)
close(conn)

# overall survival
conn <- file(series.mat.file)
open(conn)
survival <- read.table(conn, skip = 91, nrow = 1)
close(conn)

# get those lines into data.frame format
smpl.info.df <- as.data.frame(t(dplyr::bind_rows(sample.accession, 
                                                 source.name, 
                                                 tissue,
                                                 gender,
                                                 age.dx,
                                                 survival))[-1, ]) 
colnames(smpl.info.df) <- c("sample_accession", "source_name", "tissue",
                            "gender", "age_at_diagnosis_yrs", 
                            "overall_survival_yrs")

# strip extraneous strings
smpl.info.df <- smpl.info.df %>%
  dplyr::mutate(tissue = gsub("cell type: ", "", tissue),
                gender = gsub("gender: ", "", gender),
                age_at_diagnosis_yrs = gsub("age at dx \\(years\\): ", "", 
                                            age_at_diagnosis_yrs),
                overall_survival_yrs = gsub("os \\(years\\): ", "", 
                                            overall_survival_yrs))

# change "N/A" to NA
smpl.info.df[which(smpl.info.df == "N/A", arr.ind = TRUE)] <- NA
```

Write the cleaned metadata to a TSV file.

```{r}
readr::write_tsv(smpl.info.df, 
                 file.path(sample.info.dir, "GSE50021_cleaned_metadata.tsv"))
```

Clean up the workspace a bit before working with the next dataset.

```{r}
to.keep <- c("%>%", "exprs.dir", "sample.info.dir", "PrepExpressionDF")
rm(list = setdiff(ls(), to.keep))
```

### E-GEOD-26576

We used an Entrez ID BrainArray package to process this data set and we need 
gene symbols to work with PLIER.

```{r}
# SCANfast processed PCL
gse26576.file <- file.path(exprs.dir, 
                           "DIPG_E-GEOD-26576_hgu133plus2_SCANfast.pcl")

# read in the PCL file, remove the trailing "_at" added by Brainarray (these
# are Entrez gene identifiers), drop the Entrez IDs with _at appended, reordered
# such that the gene identifiers are the first column
gse26576.df <- readr::read_tsv(gse26576.file) %>%
  dplyr::mutate(EntrezID = sub("_at", "", X1)) %>%
  dplyr::select(-X1) %>%
  dplyr::select(EntrezID, dplyr::everything())
```

#### Gene identifier conversion

```{r}
# extract the Entrez ID Gene symbol mapping from org.Hs.eg.db
symbol.obj <- org.Hs.eg.db::org.Hs.egSYMBOL
mapped.genes <- AnnotationDbi::mappedkeys(symbol.obj)
symbol.list <- as.list(symbol.obj[mapped.genes])
symbol.df <- as.data.frame(cbind(names(symbol.list), unlist(symbol.list)))
colnames(symbol.df) <- c("EntrezID", "GeneSymbol")
```

Join the annotation `data.frame` to the expression `data.frame`

```{r}
annot.gse26576.df <- symbol.df %>%
  dplyr::inner_join(gse26576.df, by = "EntrezID")
rm(symbol.df)
```

Are there any duplicates?

```{r}
any(duplicated(annot.gse26576.df$GeneSymbol))
```

No, so we don't need to do anything else.
Write the `data.frame` that includes gene symbols to file.

```{r}
gse26576.output.file <- 
  file.path(exprs.dir, 
            "DIPG_E-GEOD-26576_hgu133plus2_SCANfast_with_GeneSymbol.pcl")
readr::write_tsv(annot.gse26576.df, path = gse26576.output.file)
```

#### Sample Metadata

```{r}
meta.gse26576.file <- file.path(sample.info.dir, "E-GEOD-26576.sdrf.txt")
meta.gse26576.df <- readr::read_tsv(meta.gse26576.file)
```

The sample-data relationship files from ArrayExpress (specifically, their 
column names) are pretty tidyverse-unfriendly.

```{r}
cleaned.meta.df <- data.frame(
  sample_id = gsub(" 1", "", meta.gse26576.df$`Source Name`),
  sample_file = meta.gse26576.df$`Array Data File`,
  sample_title = meta.gse26576.df$`Comment [Sample_title]`,
  age_at_diagnosis = meta.gse26576.df$`Characteristics[age at diagnosis (years)]`,
  disease_state = meta.gse26576.df$`Characteristics[disease]`,
  histology = meta.gse26576.df$`Characteristics[histology]`,
  sample_collection = meta.gse26576.df$`Characteristics[sample collection]`,
  material_type = meta.gse26576.df$`Material Type`
) %>%
  # get rid of the genomic DNA samples
  dplyr::filter(material_type == "total RNA") %>%
  dplyr::select(-material_type)
```

The `sample_file` field will match the headers of the PCL file.
Write the cleaned metadata to file.

```{r}
readr::write_tsv(cleaned.meta.df, 
                 path = file.path(sample.info.dir, 
                                  "E-GEOD-26576_cleaned_metadata.tsv"))
```

