---
title: "DIPG: Analysis"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

Analyze the DIPG `B` matrices from the MultiPLIER model

## Set up

```{r}
`%>%` <- dplyr::`%>%`
source(file.path("util", "test_LV_differences.R"))
```

Directories for this notebook

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "36")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "36")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in files

#### GSE50021

```{r}
gse50021.b.file <- file.path("results", "35", "GSE50021_recount2_B.RDS")
gse50021.b <- readRDS(gse50021.b.file)
gse50021.meta.file <- file.path("data", "sample_info", 
                                "GSE50021_cleaned_metadata.tsv")
gse50021.meta.df <- readr::read_tsv(gse50021.meta.file)
```

#### GSE26576

```{r}
gse26576.b.file <- file.path("results", "35", "E-GEOD-26576_recount2_B.RDS")
gse26576.b <- readRDS(gse26576.b.file)
gse26576.meta.file <- file.path("data", "sample_info", 
                                "E-GEOD-26576_cleaned_metadata.tsv")
gse26576.meta.df <- readr::read_tsv(gse26576.meta.file)
```

### Metadata

We should explore the metadata a little bit, because these datasets are both
quite small.
So, we'll need to sort out what kinds of analyses are possible.

```{r}
gse50021.meta.df %>%
  dplyr::group_by(tissue) %>%
  dplyr::tally()
```

```{r}
gse26576.meta.df %>%
  dplyr::group_by(disease_state) %>%
  dplyr::tally()
```

`GSE26576` does not have enough normal samples for a DIPG-normal comparison and
the other groups in `GSE26576` are not present in `GSE50021`.

We can do a comparison of DIPG-normal in `GSE50021`.

## Differential expression in `GSE50021`

```{r}
gse50021.results <- 
  LVTestWrapper(b.matrix = gse50021.b,
                sample.info.df = dplyr::mutate(gse50021.meta.df,
                                               Sample = sample_accession),
                phenotype.col = "tissue",
                file.lead = "GSE50021_normal_dipg",
                plot.dir = plot.dir,
                results.dir = results.dir)
```

```{r}
gse50021.results$limma %>%
  dplyr::filter(adj.P.Val < 0.05)
```

After revisiting [Buczkowicz, et al.](https://dx.doi.org/10.1038/ng.2936) and
the [`GSE50021`](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE50021)
accession, there is no information about which _part_ of the brain the normal
brain came from or how it was obtained. 
So, I am a bit concerned about drawing any conclusions from this analysis, as
the differences we see may not be attributable to disease state.

