---
title: "Identifying differentially expressed latent variables in 
medulloblastoma"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

Here, we're testing for latent variable differential expression between
medulloblastoma subgroups (Group 3, Group 4, and SHH).

As in `37-medulloblastoma_recount2_model`, where we prepped the data we are
analyzing here, this is using 
[Northcott, et al.](https://dx.doi.org/10.1038/nature11327)  and
[Robinson, et al.](https://dx.doi.org/10.1038/nature11213) data.

## Set up

```{r}
# pipe is required for LVTestWrapper
`%>%` <- dplyr::`%>%`
```

We have several custom functions that we've written and previously used in our 
ANCA-associated vasculitis analyses.
They are general enough that we can use them again here.

```{r}
source(file.path("util", "test_LV_differences.R"))
```

#### Directories

```{r}
# directories specific to this notebook
plot.dir <- file.path("plots", "38")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "38")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in files

We need two things to do this analysis: a B matrix (latent variable values) and
a `data.frame` that contains the sample labels for the groups that we'd like
to test for differences between.

### GSE37382 (Northcott, et al.)

```{r}
northcott.b.file <- file.path("results", "37", "GSE37382_recount2_B.RDS")
northcott.sample.file <- file.path("data", "sample_info", 
                                   "GSE37382_cleaned_metadata.tsv")
```

Read in `B` (MultiPLIER)

```{r}
northcott.b.matrix <- readRDS(northcott.b.file)
```

Read in the `data.frame` with the subgroup information and check that the
sample names match between these two files.

```{r}
northcott.sample.df <- readr::read_tsv(northcott.sample.file)
all(northcott.sample.df$source_name %in% colnames(northcott.b.matrix))
```

`LVTestWrapper` requires that the sample names/identifiers are in a column named
`Sample`, so we'll change `source_name` to `Sample`.

```{r}
colnames(northcott.sample.df)[1] <- "Sample"
```

### GSE37418 (Robinson, et al.)

We'll repeat that process for the Robinson, et al. data

```{r}
robinson.b.file <- file.path("results", "37", "GSE37418_recount2_B.RDS")
robinson.sample.file <- file.path("data", "sample_info", 
                                  "metadata_GSE37418.tsv")
```

```{r}
robinson.b.matrix <- readRDS(robinson.b.file)
robinson.sample.df <- readr::read_tsv(robinson.sample.file)
```

The subgroup information in this data set is a bit different, let's look at
the counts

```{r}
robinson.sample.df %>% 
  dplyr::group_by(subgroup) %>%
  dplyr::tally()
```

To do differential expression analysis, we'll remove the `SHH OUTLIER` and
`U` samples

```{r}
sample.index <- which(robinson.sample.df$subgroup %in% c("U",
                                                         "SHH OUTLIER"))
samples.to.remove <- robinson.sample.df$refinebio_accession_code[sample.index]
samples.to.remove
```

```{r}
remove.column.index <- which(colnames(robinson.b.matrix) %in% samples.to.remove)
```

Do the filtering

```{r}
robinson.b.matrix <- robinson.b.matrix[, -remove.column.index]
robinson.sample.df <- robinson.sample.df %>%
  dplyr::filter(!(refinebio_accession_code %in% samples.to.remove)) %>%
  dplyr::select(c("refinebio_accession_code", "subgroup", "Sex", "age",
                  "title"))
```

```{r}
all(robinson.sample.df$refinebio_accession_code %in% 
      colnames(robinson.b.matrix))
```

```{r}
colnames(robinson.sample.df)[1] <- "Sample"
```

## Test for differential expression

`LVTestWrapper` gives us 3 things: 1) differential expression results from
`limma`, 2) boxplot + jitter plots of the latent variable expression and 3)
the `B` matrix in "long" format -- this is what is used for plotting.

We'll use `"BH"` correction for multiple hypotheses testing (the default).

```{r}
northcott.results <- 
  LVTestWrapper(b.matrix = northcott.b.matrix,
                sample.info.df = northcott.sample.df,
                phenotype.col = "subgroup",
                # the boxplot output, the "long" format B data.frame (useful for 
                # plotting), and the limma differential expression results
                # will be output in files that begin with this string
                file.lead = "GSE37382_subgroup_recount2_model",
                plot.dir = plot.dir,
                results.dir = results.dir)
```

```{r}
robinson.results <- 
  LVTestWrapper(b.matrix = robinson.b.matrix,
                sample.info.df = robinson.sample.df,
                phenotype.col = "subgroup",
                file.lead = "GSE37418_subgroup_recount2_model",
                plot.dir = plot.dir,
                results.dir = results.dir)
```

### Any overlap at all?

```{r}
northcott.delvs <- 
  northcott.results$limma$LV[which(northcott.results$limma$adj.P.Val < 0.05)]
```

```{r}
robinson.delvs <- 
  robinson.results$limma$LV[which(robinson.results$limma$adj.P.Val < 0.05)]
```

```{r}
VennDiagram::venn.diagram(list(Robinson = robinson.delvs,
                               Northcott = northcott.delvs),
                          file.path(plot.dir, "Medulloblastoma_DELV_Venn.tiff"))
```

