---
title: "L2 Penalty"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

We've been asked to clarify why an L2 penalty is used on the PLIER `B`
matrix. 

We'll see what happens _in practice_ if we set the `L2` parameter to
zero.

Let's work with the data that are 500 randomly selected samples from recount2;
this is out of convenience, as the computational requirements will be 
relatively light.

The main PLIER function, for the version of PLIER we've used, begins here:
https://github.com/wgmao/PLIER/blob/a2d4a2aa343f9ed4b9b945c04326bebd31533d4d/R/Allfuncs.R#L227

## Set up

```{r}
`%>%` <- dplyr::`%>%`
library(PLIER)
```

We'll need `GetPathwayCoverage`

```{r}
source(file.path("util", "plier_util.R"))
```

#### Directories

Plot and results directories specifically for this notebook.

```{r}
# plot and result directory setup for this notebook
plot.dir <- file.path("plots", "39")
dir.create(plot.dir, recursive = TRUE, showWarnings = FALSE)
results.dir <- file.path("results", "39")
dir.create(results.dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

Working only with `n = 500` here.

```{r}
models.file <- file.path("models", "subsampled_recount2_PLIER_model_500.RDS")
models.list <- readRDS(models.file)
```

Get the prior information matrix used with the subsampling experiments.

```{r}
recount.prepped.data <- readRDS(file.path("data", "recount2_PLIER_data", 
                                          "recount_data_prep_PLIER.RDS"))
pathway.mat <- recount.prepped.data$all.paths.cm
rm(recount.prepped.data)
```

## Main

What constants were used originally (calculated from SVD)?

```{r}
lapply(models.list, function(x) x$PLIER$L1)
```

```{r}
lapply(models.list, function(x) x$PLIER$L2)
```

These are the automatically selected values.

### Training with `L2 = 0`

A custom function for setting L2 to zero.

```{r}
# plier.repeat is a list with the elements exprs (a gene expression matrix) and
# PLIER (output of PLIER); it is a repeat from script/subsampling_PLIER.R.
# use PLIER results as input to a new PLIER model, where we use the k, L1, 
# priorMat, and expression from a repeat, BUT set L2 = 0
# only intended to be used in this environment (need pathway.mat)
SetL2Zero <- function(plier.repeat) {
  results <- PLIER::PLIER(data = as.matrix(plier.repeat$exprs),
                          priorMat = pathway.mat,
                          k = nrow(plier.repeat$PLIER$B),
                          L1 = plier.repeat$PLIER$L1,
                          L2 = 0,
                          trace = TRUE)
  message("\n\n")
  return(results)
}
```

Perform the training and save the results to file (they will be small 
enough to track with Git LFS).

```{r}
penalty.zero.list <- lapply(models.list, SetL2Zero)
saveRDS(penalty.zero.list, 
        file.path(results.dir, 
                  "subsampled_recount2_PLIER_model_L2_to_zero_500.RDS"))
```

## Pathway coverage

```{r}
pathway.coverage.list <- lapply(penalty.zero.list, GetPathwayCoverage)
```

```{r}
coverage.df <- reshape2::melt(lapply(pathway.coverage.list, 
                                     function(x) x$pathway))
colnames(coverage.df) <- c("L2_zero_value", "seed")
```

Read in the results from the original subsampling experiment

```{r}
coverage.file <- file.path("results", "30", "subsampled_pathway_coverage.tsv")
subsampling.coverage.df <- readr::read_tsv(coverage.file) %>%
  dplyr::filter(metric == "pathway coverage",
                sample_size == "500") %>%
  dplyr::select(value, seed)
colnames(subsampling.coverage.df)[1] <- "automatic_value"
```

Join and find the difference in pathway coverage between the "normal" model and 
the `L2 = 0` model

```{r}
coverage.df <- coverage.df %>%
  dplyr::mutate(seed = as.integer(seed)) %>%
  dplyr::inner_join(subsampling.coverage.df, by = "seed") %>%
  dplyr::select(seed, dplyr::everything()) %>%
  dplyr::mutate(difference = automatic_value - L2_zero_value)
```

```{r}
coverage.df
```

```{r}
readr::write_tsv(coverage.df, 
                 file.path(results.dir, "pathway_coverage_with_difference.tsv"))
```

### LV-pathway association summaries

Let's look at an example summary data.frame for `L2 = 0`

```{r}
penalty.zero.list$`2876`$summary %>% 
  dplyr::filter(FDR < 0.05)
```

And another:

```{r}
penalty.zero.list$`8828`$summary %>% 
  dplyr::filter(FDR < 0.05)
```

By going through the pages of these summaries, we can see that gene sets
related to the spliceosome are woefully overrepresented.

### `Z` matrix

One explanation for this is that the loadings, `Z`, are no longer capturing
sparse combinations of pathways.
We can check the number of positive entries for each latent variable (column).

```{r}
example.z.matrix <- penalty.zero.list$`2876`$Z
positive.counts <- apply(example.z.matrix, 2, function(x) sum(x > 0))
```

Compare to the `Z` matrix for automatically selected `L2` (same seed and 
therefore expression matrix)

```{r}
auto.example.z.matrix <- models.list$`2876`$PLIER$Z
auto.positive.counts <- apply(auto.example.z.matrix, 2, function(x) sum(x > 0))
```

A `data.frame` for plotting

```{r}
z.positive.df <- data.frame(
  positive_count = c(positive.counts, auto.positive.counts),
  model_type <- c(rep("L2 = 0", length(positive.counts)),
                  rep("L2 automatic", length(auto.positive.counts)))
)
```

Density plot of positive values

```{r}
z.positive.df %>%
  ggplot2::ggplot(ggplot2::aes(x = positive_count, group = model_type,
                               fill = model_type)) +
  ggplot2::geom_density(alpha = 0.5) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "number of positive entries in Z",
                title = "Effect of L2 = 0 on Loadings") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
                 text = ggplot2::element_text(size = 12)) +
  ggplot2::guides(fill = ggplot2::guide_legend(title = "parameter")) +
  ggplot2::scale_fill_manual(values = c("#FFFFFF", "#545454"))
```

```{r}
ggplot2::ggsave(file.path(plot.dir, 
                          "L2_effect_on_Z_positive_entries_density.pdf"),
                plot = ggplot2::last_plot())
```

