---
title: "Display items: Medulloblastoma"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

We've looked into differential expression of latent variables in the
[Northcott, et al.](https://doi.org/10.1038/nature11327) 
and [Robinson, et al.](https://doi.org/10.1038/nature11213) 
medulloblastoma datasets. 
Specifically, we looked at differential expression between subgroups of 
patients.

## Set up

#### Functions and libraries

```{r}
`%>%` <- dplyr::`%>%`
library(ggplot2)
```

#### Directories

```{r setup}
# set directory to top directory of project
knitr::opts_knit$set(root.dir = "..")
```

```{r}
plot.dir <- file.path("figure_notebooks", "figures")
dir.create(plot.dir, showWarnings = FALSE, recursive = TRUE)
```

## Read in data

```{r}
northcott.file <- 
  file.path("results", "38", 
            "GSE37382_subgroup_recount2_model_B_long_sample_info.tsv")
northcott.df <- readr::read_tsv(northcott.file)
```

```{r}
robinson.file <- 
  file.path("results", "38",
            "GSE37418_subgroup_recount2_model_B_long_sample_info.tsv")
robinson.df <- readr::read_tsv(robinson.file)
```

## Data wrangling

```{r}
unique(robinson.df$subgroup)
```

```{r}
unique(northcott.df$subgroup)
```

Recode the `subgroup` labels in Robinson to match Northcott and change some
column names to be more consistent as well

```{r}
robinson.df <- robinson.df %>%
  dplyr::mutate(subgroup = dplyr::case_when(
    subgroup == "G4" ~ "Group 4",
    subgroup == "G3" ~ "Group 3",
    TRUE ~ subgroup
  ), sex = Sex) %>%
  dplyr::select(-Sex)
unique(robinson.df$subgroup)
```

We'll join these two `data.frame` together to facilitate plotting, filtering 
for latent variables of interest after the fact.

```{r}
selected.columns <- c("LV", "Sample", "Value", "subgroup")
lv.df <- dplyr::bind_rows(
  dplyr::select(northcott.df, !!rlang::enquo(selected.columns)),
  dplyr::select(robinson.df, !!rlang::enquo(selected.columns)),
  .id = "dataset"
) %>%
  dplyr::mutate(dataset = dplyr::case_when(
    dataset == 1 ~ "Northcott",
    dataset == 2 ~ "Robinson"
  ))
```

A decent number of the top "named" LVs that are differentially expressed 
(between subgroups) in Northcott, et al. are related to translation, RNA
processing, ribosomes, etc., so we'll plot some of these in both the Northcott,
et al. and Robinson, et al. datasets.

```{r}
lv.df %>%
  dplyr::filter(LV %in% c("161,REACTOME_TRNA_AMINOACYLATION",
                          "707,REACTOME_PEPTIDE_CHAIN_ELONGATION")) %>%
  dplyr::mutate(LV = gsub(",", ", ", gsub("_", " ", LV))) %>%
  dplyr::mutate(LV = paste("LV", LV)) %>%
  ggplot(aes(x = subgroup, group = subgroup, y = Value, color = subgroup)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, width = 0.3) +
  facet_wrap(LV ~ dataset, scales = "free_x") +
  labs(title = "Medulloblastoma Subgroups", y = "LV expression value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none") +
  scale_color_manual(values = c("#000000", "#E69F00", "#56B4E9", "#009E73"))
```

```{r}
ggsave(file.path(plot.dir, "medulloblastoma_G4_LV161_LV707.pdf"))
```

