---
title: "Tweaking display items: SLE U plot, AUC cut points"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni 2018**

## Set up 

```{r setup}
knitr::opts_knit$set(root.dir = "..")
```

```{r}
library(PLIER)
`%>%` <- dplyr::`%>%`
```

```{r}
plot.dir <- file.path("figure_notebooks", "figures")
# color palette for PCA plots
plot.color.pal <- c("#54FF9F", "#43CD80", "#2E8B57", "#006400", "#FF8C00",
                    "#8B4500", "#000080")
```

## Read in data

PLIER model

```{r}
# read in SLE WB PLIER model
plier.file <- file.path("results", "05", "SLE-WB_PLIER_model.RDS")
sle.plier <- readRDS(plier.file)
```

Sample-dataset mapping file

```{r}
# read sample dataset mapping file
sd.file <- file.path("data", "sample_info", "sle-wb_sample_dataset_mapping.tsv")
sd.df <- readr::read_tsv(sd.file)
```

## Plotting

### `U` plot with selected latent variables

We supplied column indices for LVs that we focus on in the main text of the 
paper.

```{r}
pdf(file.path(plot.dir, "SLE_U_auc0.75_custom_col_index.pdf"))
PLIER::plotU(sle.plier, auc.cutoff = 0.75, fontsize_row = 7,
             fontsize_col = 7, indexCol = c(6, 69, 110, 87, 34, 2, 27, 122, 86,
                                            25, 60, 52, 59, 109, 72, 8, 74))
dev.off()
```

### AUC

We picked `AUC > 0.75` as a cutoff for the "pathway-associated" latent variables
in our SLE WB PLIER overview figure, but I can not recall how much we dug into
what those AUC values look like overall.
We'll explore this here, rather than `05-sle-wb-PLIER`, so we don't have to
retrain the model.

_Note: These AUC values are only calculated for LV with non-zero values in `U`._

**What does the distribution of AUC values look like?**

```{r}
sle.plier$summary %>%
  ggplot2::ggplot(ggplot2::aes(x = AUC)) +
  ggplot2::geom_density(fill = "#000000", alpha = 0.2) +
  ggplot2::theme_bw() +
  ggplot2::ggtitle("SLE WB Model LV AUC Values") +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
                 text = ggplot2::element_text(size = 15))
```
```{r}
summary(sle.plier$summary$AUC)
```

We're a bit above the 3rd quartile at `0.75`. What are the 90th and 95th 
percentiles?

```{r}
quantile(sle.plier$summary$AUC, 0.9)
```

```{r}
quantile(sle.plier$summary$AUC, 0.95)
```

```{r}
# this function is only intended to be run in this notebook/context!
PlotPCAWrapper <- function(plier.result, sd.df, color.palette, 
                           cutoff = 0.75) {
  # plier.result: output of PLIER::PLIER
  # sd.df: the sample-dataset mapping data.frame, contains a column named 
  #        'SampleID', these sample ids need to match the column names of
  #        the B matrix or this will not work!
  # color.palette: color palette to be used with plots, this should be of 
  #                length(number of datasets) or greater
  # cutoff: what percentile should be used for the plots
  
  # since we'll produce _two_ plots for the cutoff -- those LVs above the 
  # threshold and those LVs with AUC values less than or equal to the cutoff
  # let's functionalize that
  PCAtoDataFrame <- function(filt.b.matrix){
    # PCA 
    lv.pc <- prcomp(t(filt.b.matrix))
    cum.var.exp <- cumsum(lv.pc$sdev^2 / sum(lv.pc$sdev^2))
    
    # PC1-2 in form suitable for ggplot2
    lv.df <- as.data.frame(cbind(rownames(lv.pc$x),
                                 lv.pc$x[, 1:2]))
    colnames(lv.df)[1] <- "SampleID"
    lv.df <- dplyr::full_join(lv.df, sd.df, by = "SampleID") %>%
      dplyr::mutate(PC1 = as.numeric(as.character(PC1)),
                    PC2 = as.numeric(as.character(PC2)),
                    Dataset = factor(Dataset, 
                                     levels = c("E-GEOD-39088",
                                                "E-GEOD-61635",
                                                "E-GEOD-72747",
                                                "E-GEOD-11907",
                                                "E-GEOD-49454",
                                                "E-GEOD-65391",
                                                "E-GEOD-78193")))
    
    return(list(lv.df = lv.df, cum.var.exp = cum.var.exp))
  }
  
  # wrapper for the plots themselves since there are two!
  PlotScatterplot <- function(lv.df, color.pal, plot.title,
                              cum.var.exp) {
    # where lv.df and cum.var.exp are the output of PCAtoDataFrame
    ggplot2::ggplot(lv.df, ggplot2::aes(x = PC1, y = PC2, 
                                        colour = Dataset)) +
      ggplot2::geom_point(alpha = 0.5) +
      ggplot2::scale_color_manual(values = color.pal) +
      ggplot2::theme_bw() +
      ggplot2::labs(x = paste0("PC1 (cum var exp = ", 
                               round(cum.var.exp[1], 3), ")"),
                    y = paste0("PC2 (cum var exp = ", 
                               round(cum.var.exp[2], 3), ")"), 
                    title = plot.title) +
      ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, 
                                                        face = "bold"),
                     text = ggplot2::element_text(size = 12),
                     legend.position = "none")
  }
  
  # using the user-specified percentile, identify the LVs with values higher
  # than that cutoff 
  auc.cutoff.value <- quantile(plier.result$summary$AUC, cutoff)
  lvs.to.keep <- unique(plier.result$summary$`LV index`[
    which(plier.result$summary$AUC > auc.cutoff.value)
  ])
  
  # filter the B matrix to only the LVs above the threshold + plot
  keep.b.matrix <- plier.result$B[as.integer(lvs.to.keep), ]
  keep.lv.df <- PCAtoDataFrame(filt.b.matrix = keep.b.matrix)
  keep.plot <- PlotScatterplot(keep.lv.df$lv.df, 
                               color.pal = color.palette,
                               plot.title = paste("Latent Variables AUC >",
                                                  round(auc.cutoff.value, 3)),
                               cum.var.exp = keep.lv.df$cum.var.exp)
  
  # other LVs
  other.b.matrix <- plier.result$B[-as.integer(lvs.to.keep), ]
  other.lv.df <- PCAtoDataFrame(filt.b.matrix = other.b.matrix)
  other.plot <- PlotScatterplot(other.lv.df$lv.df, 
                                color.pal = color.palette,
                                plot.title = paste("Latent Variables AUC <=",
                                                   round(auc.cutoff.value, 3)),
                                cum.var.exp = other.lv.df$cum.var.exp)
  
  # return a list of two plots
  return(list(above.plot = keep.plot, other.plot = other.plot))   
}
```

```{r}
cutoff.values <- c(0.75, 0.9, 0.95)
plot.lists <- lapply(cutoff.values,
                     function(x) {
                       PlotPCAWrapper(plier.result = sle.plier,
                                      sd.df = sd.df,
                                      color.palette = plot.color.pal,
                                      cutoff = x)
                     })
```

### Multipanel plot

We'll start the panel labeling with **B** because panel **A** will be the
pairs plot from PCA on the gene-level expression values from this compendium.

```{r}
# wrapper for the two panels for each percentile cut off
TwoPanelWrapper <- function(plot.list, panel.labels = c("B", "C"),
                            side.label = "75th percentile") {
  two.panel <- cowplot::plot_grid(plotlist = plot.list, 
                                  align = "h", ncol = 2, scale = 0.95,
                                  labels = panel.labels)
  plot.label <- cowplot::ggdraw() + 
    cowplot::draw_label(side.label, fontface = "bold", size = 20, 
                        angle = -90)
  return(cowplot::plot_grid(two.panel, plot.label, align = "h", ncol = 2, 
                            labels = NULL, rel_widths = c(1, 0.1)))
}
```

75th percentile

```{r}
seventy.fifth <- TwoPanelWrapper(plot.list = plot.lists[[1]],
                                 panel.labels = c("B", "C"),
                                 side.label = "75th percentile")
seventy.fifth
```

90th percentile

```{r}
ninetieth <- TwoPanelWrapper(plot.list = plot.lists[[2]], 
                             panel.labels = c("D", "E"),
                             side.label = "90th percentile")
```

95th percentile

```{r}
ninety.fifth <- TwoPanelWrapper(plot.list = plot.lists[[3]], 
                                panel.labels = c("F", "G"),
                                side.label = "95th percentile")
```

Final plot! 

```{r}
pdf(file.path(plot.dir, "SLE_PCA_different_cutpoints.pdf"), width = 8.5,
    height = 11)
cowplot::plot_grid(plotlist = list(seventy.fifth, ninetieth, ninety.fifth), 
                   ncol = 1, nrow = 3)
dev.off()
```

